%=====================================================================
\section{The Unique Dimension: Why $d = 5$}
\label{ch:whyd5}
%=====================================================================

Chapter~\ref{ch:whyC} established that points carry vectors $\psi_i \in \CC^d$ for some $d$. We now prove that $d = 5$ is the unique value producing non-trivial physics. The argument proceeds in four steps: (i) identify the ``atomic'' dimensions, (ii) show that repeated atoms annihilate via swap symmetry, (iii) show that non-atomic dimensions decompose into atoms, and (iv) verify that only $d = 2 + 3 = 5$ survives.

\subsection{Atomic dimensions}

\begin{definition}[Atomic dimension]
An integer $n \geq 2$ is \emph{atomic} if it cannot be written as $n = a + b$ with $a, b \geq 2$. The integer $n = 1$ is excluded on purely algebraic grounds: $\text{SU}(1) = \{1\}$ is the trivial group, with $1^2 - 1 = 0$ generators. A sector with $n = 1$ has no gauge bosons, no force, and no internal degrees of freedom. It cannot constitute an independent gauge sector. (The U(1) electromagnetic factor arises as the \emph{relative phase between two sectors with $n \geq 2$}, not as a sector in its own right; see Sec.~\ref{sec:glue}.) This exclusion requires no observational input --- it is the statement that the trivial group is trivial.
\end{definition}

\begin{lemma}[The atoms are $\{2, 3\}$]
\label{lem:atoms}
The only atomic dimensions are $2$ and $3$.
\end{lemma}

\begin{proof}
For $n \geq 4$: write $n = 2 + (n - 2)$, where $n - 2 \geq 2$. Hence $n$ is not atomic.

For $n = 2$: the only partition into parts $\geq 1$ is $2 = 1 + 1$, but $1 < 2$. No valid decomposition exists. Atomic.

For $n = 3$: the partitions are $3 = 1 + 2$ and $3 = 1 + 1 + 1$. Since $1 < 2$, no valid decomposition exists. Atomic.
\end{proof}

\begin{remark}
Every integer $n \geq 2$ can be written as $n = 2a + 3b$ for non-negative integers $a, b$. The atoms $\{2, 3\}$ are the additive generators of all dimensions above the ``phase-only'' threshold.
\end{remark}

\subsection{Swap elimination: Group-theoretic proof}

\begin{theorem}[Swap Automorphism]
\label{thm:swap}
Let $\CC^d = \CC^a \oplus \CC^a$ with $a \geq 2$. The gauge group $G = \text{SU}(a) \times \text{SU}(a) \times \text{U}(1)$ admits an outer automorphism $\sigma$ that forces the vacuum to be symmetric. Symmetric vacua produce no physical structure.
\end{theorem}

\begin{proof}
\textbf{Step 1: The automorphism.}
Define the swap:
\begin{equation}
\sigma: (g_1, g_2, e^{i\theta}) \;\mapsto\; (g_2, g_1, e^{-i\theta}).
\end{equation}
Since $\text{SU}(a)_1 \cong \text{SU}(a)_2$, this is a well-defined outer automorphism with $\sigma^2 = \text{id}$, generating $\mathbb{Z}_2$.

\textbf{Step 2: Vacuum classification.}
Let $|\Omega\rangle$ be a vacuum state.

\emph{Case (a): $\sigma|\Omega\rangle = |\Omega\rangle$.}
The two sectors are physically identical. All observables are $\sigma$-invariant: if $\mathcal{O}$ is any physical quantity, $\sigma(\mathcal{O}) = \mathcal{O}$. This means no distinction between the two sectors is possible --- no chirality ($\sigma$ maps left to right), no CP violation ($\sigma$ maps matter to antimatter), no force hierarchy (both sectors have identical coupling). Without chiral fermions, the Standard Model structure cannot arise. (This has been verified numerically in lattice simulations, but the conclusion follows from $\sigma$-invariance of the representation ring of $\text{SU}(a) \times \text{SU}(a)$ alone.)

\emph{Case (b): $\sigma|\Omega\rangle \neq |\Omega\rangle$.}
Spontaneous $\mathbb{Z}_2$ breaking. Domain walls form between $|\Omega\rangle$ and $\sigma|\Omega\rangle$. Tunneling restores symmetry in the ensemble average. The long-range physics is again symmetric.

In both cases, no net asymmetry survives. Physics is trivial.
\end{proof}

\begin{corollary}[Repeated atoms annihilate]
\label{cor:annihilate}
Any decomposition $d = n_1 + n_2 + \cdots + n_k$ containing a repeated pair $n_i = n_j$ admits a swap $\sigma_{ij}$ on that pair. By Theorem~\ref{thm:swap}, the pair annihilates. The effective dimension reduces by $2n_i$.
\end{corollary}

\begin{example}
$d = 4 = \mathbf{2+2}$: swap $\to$ trivial. \;
$d = 6 = \mathbf{3+3}$: swap $\to$ trivial. \;
$d = 7 = \mathbf{2+2}+3$: the 2-pair annihilates, leaving only 3. \;
$d = 8 = 2+\mathbf{3+3}$: the 3-pair annihilates, leaving only 2. \;
$d = 10 = \mathbf{2+2}+\mathbf{3+3}$: both pairs annihilate $\to$ nothing.
\end{example}

\subsection{Swap elimination: Topological proof}

\begin{theorem}[Grassmannian Attractor]
\label{thm:grassmann}
If $a = b$, the Grassmannian $\text{Gr}(a, 2a)$ of causal decompositions has a $\mathbb{Z}_2$-invariant fixed locus that is a topological attractor, trapping the dynamics in a structureless configuration.
\end{theorem}

\begin{proof}
\textbf{Step 1.}
The space of decompositions is $\mathcal{G} = \text{Gr}(a, d) = \text{U}(d)/(\text{U}(a) \times \text{U}(b))$.

\textbf{Step 2.}
When $a = b$: the complement map $\tau: V_S \mapsto V_S^\perp$ is a well-defined involution on $\mathcal{G}$ (since $\dim V_S^\perp = b = a$). For $a \neq b$: $\tau$ maps $\text{Gr}(a, d) \to \text{Gr}(b, d) \neq \text{Gr}(a, d)$. No involution.

\textbf{Step 3.}
The fixed locus $\mathcal{G}^\tau = \{V_S : V_S = V_S^\perp\} = \text{U}(a)/\text{O}(a)$ is non-empty for all $a \geq 1$.

\textbf{Step 4.}
Entropy maximization drives the system toward $\mathcal{G}^\tau$ (the maximally disordered symmetric configuration). Once there, $V_S \cong V_T$, freezing is symmetric, $\hbar$ is constant. No physics.

For $a \neq b$: no involution $\Rightarrow$ no attractor $\Rightarrow$ asymmetric dynamics is forced.
\end{proof}

\subsection{Gravity requires $d \geq 5$}

\begin{proposition}
For $d_\text{spacetime} = d - 1 \leq 3$, gravity has zero propagating degrees of freedom.
\end{proposition}

\begin{proof}
Graviton polarizations: $n_\text{grav} = d_\text{spacetime}(d_\text{spacetime} - 3)/2$. For $d_\text{spacetime} \leq 3$: $n_\text{grav} \leq 0$. No gravitational waves, no black hole horizons. The Gram determinant $\det(G_h)$ encodes no dynamical information: spacetime curvature is fully determined by local matter content.
\end{proof}

\subsection{The uniqueness theorem}

\begin{theorem}[Uniqueness of $(2,3)$ --- independent of $N$]
\label{thm:d5}
For any integer $N \geq 5$, the unique atomic decomposition of $\CC^N$ supporting (a) chiral fermions, (b) CP violation, (c) connected spacetime, and (d) charge quantization, is $(n_S, n_T) = (3, 2)$, yielding $\text{SU}(3) \times \text{SU}(2) \times \text{U}(1)$.
\end{theorem}

\begin{proof}
(1) Atoms $= \{2, 3\}$ (Lemma~\ref{lem:atoms}). (2) Any $N \geq 2$ decomposes as $N = 2a + 3b$. (3) Repeated atoms are trivial: the swap $\sigma: \text{SU}(n)_1 \times \text{SU}(n)_2 \to \text{SU}(n)_2 \times \text{SU}(n)_1$ forces vector-like representations, precluding CP violation. (4) At most one copy of each atom survives. (5) $\{2\}$ alone: $n_S = 0$, no space. $\{3\}$ alone: $n_T = 0$, no time. Both have $\pi_1 = 0$: no charge quantization. (6) $\text{U}(1)$ is the unique glue: $\pi_1(\text{U}(1)) = \mathbb{Z}$. (7) Steps 1--6 depend only on the algebra of $\text{SU}(2)$, $\text{SU}(3)$, $\text{U}(1)$, independent of $N$.
\end{proof}

\begin{corollary}
The question ``why $d = 5$?'' dissolves. The physical content is $(2, 3)$ regardless of $d$. The number $5 = 2 + 3$ is a consequence, not a cause. Whether $N = 5$ or $N = 10^{60}$, the physically non-trivial content is $(2, 3)$.
\end{corollary}

\begin{remark}[Spectral invisibility]
Swap annihilation creates no spectral gap at rank 5. The chiral eigenvalues $\lambda_1, \ldots, \lambda_5$ and the trivial eigenvalues $\lambda_6, \ldots, \lambda_{d_\text{indep}}$ merge into a continuous spectrum as $N \to \infty$ (gap $\sim 1/\sqrt{N}$). The true spectral gap is at rank $d_\text{indep}$ (independent dimensions), where $\lambda_{d_\text{indep}+1} = 0$ exactly. The distinction between chiral and trivial dimensions is purely representation-theoretic: physics is algebraic, not spectral.
\end{remark}

\subsection{The cosmic attractor}

The uniqueness theorem has a dynamical interpretation. Starting from $\text{SU}(N)$ for any $N$:
\begin{equation}
N = \underbrace{2 + 2 + \cdots + 2}_{a\text{ pairs}} + \underbrace{3 + 3 + \cdots + 3}_{b\text{ pairs}} + \text{remainder}
\end{equation}

All repeated atoms annihilate by Corollary~\ref{cor:annihilate}. At most one 2 and one 3 survive:
\begin{equation}
\text{SU}(N) \;\xrightarrow{\;\text{swap annihilation}\;}\; \text{SU}(3) \times \text{SU}(2) \times \text{U}(1)
\end{equation}
\emph{independent of $N$}. The Standard Model gauge group is the universal attractor of dimension reduction.

\begin{remark}[Why not fewer factors?]
No proper subset of $\text{SU}(3) \times \text{SU}(2) \times \text{U}(1)$ is viable:
\begin{itemize}
\item $\text{SU}(3)$ alone: $d = 3$, so $n_T = 0$. No temporal sector, no unitary evolution, no time. Dead.
\item $\text{SU}(2)$ alone: $d = 2$, so $n_S = 0$. No spatial sector, no matter, no extent. Dead.
\item $\text{SU}(3) \times \text{SU}(2)$ without $\text{U}(1)$: the two sectors are disconnected --- no relative phase, no electromagnetic coupling. Each sector independently faces the problems above ($\CC^3$ has no time; $\CC^2$ has no space). Without the glue, the product decays into its factors, and each factor dies.
\end{itemize}
$\text{SU}(3) \times \text{SU}(2) \times \text{U}(1)$ is the \emph{minimum stable configuration}: $\text{SU}(3)$ provides space, $\text{SU}(2)$ provides time, and $\text{U}(1)$ prevents their separation. Removing any factor collapses the structure.
\end{remark}

\subsection{The U(1) glue}
\label{sec:glue}

After the split $\CC^5 = \CC^3 \oplus \CC^2$, the U(1) factor is the relative phase between the two sectors --- the unique element of $\CC$ that Chapter~\ref{ch:whyC} established. The tracelessness of hypercharge:
\begin{equation}
n_S \cdot y_\text{color} + n_T \cdot y_\text{weak} = 0 \qquad\Rightarrow\qquad \frac{y_\text{color}}{y_\text{weak}} = -\frac{n_T}{n_S} = -\frac{2}{3}
\end{equation}
enforces exact charge conservation: what one sector gives, the other takes. The electromagnetic force is not an accident but the \emph{necessary glue} connecting space and time.

\subsection{Summary}

\begin{center}
\begin{tabular}{cccl}
\hline
$d$ & Decomposition & Physics? & Reason \\
\hline
$\leq 4$ & --- & No & Gravity: 0 degrees of freedom \\
4 & $2+2$ & No & Swap $\to$ symmetric $\to$ trivial \\
\textbf{5} & $\mathbf{2+3}$ & \textbf{Yes} & \textbf{Unique distinct atomic pair} \\
6 & $3+3$ & No & Swap $\to$ trivial \\
7 & $2+2+3$ & No & Repeated 2's $\to$ annihilate \\
$\geq 6$ & Any & No & All repeated atoms annihilate $\to$ only $(3,2)$ \\
$N$ & Any & $\to \mathbf{5}$ & Universal attractor \\
\hline
\end{tabular}
\end{center}

\noindent The spacetime dimension $d_\text{spacetime} = 4$ is a theorem, not a parameter.
