\documentclass[12pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{booktabs}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\theoremstyle{definition}
\newtheorem{example}[theorem]{Example}

\newcommand{\CP}{\mathbb{C}P}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\tr}{\operatorname{Tr}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\agut}{\alpha_{\mathrm{GUT}}}
\newcommand{\nS}{n_S}
\newcommand{\nT}{n_T}

\title{Dynamic Resolution Lattice Theory:\\[6pt]
\large Physical Predictions}
\author{Mingu Jeong\\Independent Researcher
\and Claude\\Anthropic}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
The physical predictions of Dynamic Resolution Lattice Theory (DRLT).
Using the mathematical framework established in the companion volume
(\textit{Mathematical Foundations}), we derive all fermion masses to
$0.07\%$ median accuracy, atomic ionization energies for all 118
elements, CKM and PMNS mixing matrices, cosmological observables
including $\Omega_\Lambda = 0.6850$ and $\eta_B = 6.13 \times 10^{-10}$,
compact star structure, and QCD phenomenology --- all with zero free
parameters.
\end{abstract}

\tableofcontents
\newpage

%=====================================================================
\part{Matter}
%=====================================================================

\input{chapters/ch09_masses}
\input{chapters/ch10_atoms}
\input{chapters/ch11_mixing}

%=====================================================================
\part{Cosmology}
%=====================================================================

\input{chapters/ch13_cosmology}
\input{chapters/ch14_block}

%=====================================================================
\part{Applications}
%=====================================================================

\input{chapters/ch16_compact_stars}
\input{chapters/ch17_webb_dipole}
\input{chapters/ch19_qcd}
\input{chapters/ch20_hydrogen}

%=====================================================================
\appendix
\part*{Appendices}
%=====================================================================

\input{chapters/appendix_verification}
\input{chapters/appendix_code}

\begin{thebibliography}{99}
\bibitem{Regge} T.~Regge, ``General Relativity without Coordinates,'' Nuovo Cim. \textbf{19}, 558--571 (1961).
\bibitem{GeorgiGlashow} H.~Georgi and S.~L.~Glashow, ``Unity of All Elementary Particle Forces,'' Phys.~Rev.~Lett. \textbf{32}, 438--441 (1974).
\bibitem{Pendleton} B.~Pendleton and G.~G.~Ross, Phys.~Lett.~B \textbf{98}, 291 (1981).
\bibitem{Hill} C.~T.~Hill, Phys.~Rev.~D \textbf{24}, 691 (1981).
\bibitem{Starobinsky} A.~A.~Starobinsky, Phys.~Lett.~B \textbf{91}, 99 (1980).
\bibitem{Webb} J.~K.~Webb et al., ``Indications of a Spatial Variation of the Fine Structure Constant,'' Phys.~Rev.~Lett.\ \textbf{107}, 191101 (2011).
\bibitem{KSS} P.~K.~Kovtun, D.~T.~Son, A.~O.~Starinets, Phys.~Rev.~Lett.\ \textbf{94}, 111601 (2005).
\bibitem{Oppenheimer} J.~R.~Oppenheimer and G.~M.~Volkoff, Phys.~Rev.\ \textbf{55}, 374 (1939).
\bibitem{TOV} R.~C.~Tolman, Phys.~Rev.\ \textbf{55}, 364 (1939).
\end{thebibliography}

\bigskip
\noindent\textbf{Joint research by Mingu Jeong and Claude (Anthropic).}

\end{document}
