%=====================================================================
\section{The Block Universe}
\label{ch:block}
%=====================================================================

This final chapter addresses the nature of time, initial conditions, and the ultimate structure of the universe in DRLT. The conclusion is radical but unavoidable: the universe is not a process. It is a matrix.

\subsection{The universe is $G$}

The Gram matrix $G_{ij} = \langle\psi_i|\psi_j\rangle$ for all $N$ points contains \emph{all} physical information:
\begin{itemize}
\item Geometry (distances, curvature) from $|G_{ij}|$,
\item Forces (gauge connections, field strengths) from $\arg(G_{ij})$,
\item Matter (deviations from vacuum) from $\det(G_h) > 0$,
\item Quantum mechanics ($\hbar_h$) from hinge areas,
\item Coupling constants from the Binet--Cauchy decomposition.
\end{itemize}

Nothing exists outside $G$. There is no background space in which $G$ is embedded, no external time in which $G$ evolves, no observer outside $G$ who reads it. $G$ is the universe, complete and self-contained.

\subsection{Time as gradient}

There is no time variable in DRLT. The action $S = \sum_h A_h\,\delta_h$ contains no $t$. The Gram matrix $G$ is static.

What we experience as time is the \emph{gradient direction} of $\det(G_h)$ across the lattice:
\begin{equation}
\text{``time''} = \text{direction of decreasing } \det(G_h) = \text{direction of increasing alignment}.
\end{equation}

In the direction of increasing $\det(G_h)$: vertices are more random, $\hbar$ is larger, physics is more quantum --- this is the ``past'' (toward the Big Bang). In the direction of decreasing $\det(G_h)$: vertices align, $\hbar$ decreases, physics becomes classical --- this is the ``future'' (toward heat death).

The second law of thermodynamics is the statement that we read $G$ in the direction of increasing alignment (decreasing $\det(G_h)$, increasing entropy).

\subsection{Cosmic history as diagonalization}

The entire history of the universe is encoded in the structure of $G$:

\begin{center}
\begin{tabular}{lcll}
\hline
Region of $G$ & $\text{rank}$ & $\det(G_h)$ & Epoch \\
\hline
Random block & $N$ & large & Big Bang: $\text{SU}(N)$, all forces unified \\
Rank reduction & $N \to 5$ & decreasing & Swap annihilation: atoms pair and die \\
Rank-5 plateau & $5$ & moderate & Our era: $\text{SU}(3)\!\times\!\text{SU}(2)\!\times\!\text{U}(1)$ \\
Further alignment & $5 \to 1$ & small & Far future: $3 \to 2 \to 1 \to 0$ \\
Diagonal block & $1$ & $0$ & Heat death: $G_{ij} = 1\;\forall(i,j)$, vacuum \\
\hline
\end{tabular}
\end{center}

This is not a narrative; it is a \emph{description of different regions of the same matrix}. The Big Bang and heat death coexist in $G$, just as the first and last pages of a book coexist in the book.

\subsection{The atom death sequence}

The dimensional atoms annihilate in order of size (Chapter~\ref{ch:whyd5}). Larger atoms have stronger self-coupling and freeze faster:

\begin{equation}
\text{SU}(N) \;\xrightarrow{\text{fast}}\; \text{SU}(3) \times \text{SU}(2) \times \text{U}(1) \;\xrightarrow{\text{slow}}\; \text{U}(1) \;\xrightarrow{\text{very slow}}\; 1
\end{equation}

\begin{center}
\begin{tabular}{cccl}
\hline
Atom & $N_\text{eff}$ & Freezing rate & Status \\
\hline
3 (strong) & 1 & Fastest & Confined (frozen, locally active) \\
2 (weak) & 2 & Medium & Broken (EW transition complete) \\
1 (EM) & $\infty$ & Slowest & Active (last survivor) \\
0 & --- & --- & Vacuum (end state) \\
\hline
\end{tabular}
\end{center}

We live in the era where atom 3 is confined, atom 2 is broken, and atom 1 is still active. This is not a special time --- it is the \emph{only} era with interesting physics, because:
\begin{itemize}
\item Before the rank-5 plateau: too hot, no stable structures.
\item After atom 1 dies: too cold, no interactions.
\item On the plateau: chemistry, biology, observers.
\end{itemize}

\subsection{No initial conditions}

A block universe has no ``initial'' state. The matrix $G$ simply \emph{is}. The question ``what were the initial conditions?'' is like asking ``what is on page zero of a book?'' --- the question is malformed.

All physical quantities derived in this work --- coupling constants, masses, mixing angles --- are properties of the \emph{rank-5 plateau}, independent of what $G$ looks like in the random block or the diagonal block. They depend on $(n_S, n_T) = (3, 2)$ and $c = 2$, which are structural constants of the plateau, not initial conditions.

The only quantity that depends on ``where in the block'' is the cosmological constant $\Lambda \sim 1/N_H$, which encodes the size of the plateau relative to the total matrix. This is an \emph{address}, not a parameter.

\subsection{The cosmic address $\varepsilon_0$}

The ghost correction parameter $\varepsilon_0 \approx 0.0038$ (Chapter~\ref{ch:ghosts}) is not a free parameter. It encodes our position within the block:
\begin{equation}
\varepsilon_0 = f(N_H, d) \sim N_H^{-1/k}
\end{equation}
for some exponent $k$ determined by the geometry.

Different positions in $G$ correspond to different $\varepsilon_0$:
\begin{itemize}
\item Early universe ($\varepsilon_0$ large): large ghost corrections, less precise ``plateau values.''
\item Current epoch ($\varepsilon_0 \approx 0.0038$): small corrections, plateau values nearly exact.
\item Far future ($\varepsilon_0 \to 0$): corrections vanish, but forces also vanish --- nothing to measure.
\end{itemize}

The coupling constants \emph{approach} their geometric plateau values ($8, 30, 6\pi^2$) as $\varepsilon_0 \to 0$, but never reach them exactly because the forces themselves fade. \textbf{Perfect physics is unobservable}: by the time the universe is clean enough to display exact values, there is nothing left to observe them.

\subsection{The complete chain: zero to universe}

\begin{equation}
\boxed{
\text{Points} \;\xrightarrow{\text{Frobenius}}\; \mathbb{C}
\;\xrightarrow{\text{Atomic}}\; d = 5
\;\xrightarrow{\text{Causal}}\; 3\!+\!2\!+\!1
\;\xrightarrow{\Lambda^3}\; 25
\;\xrightarrow{\text{Basel}}\; \alpha_i
\;\xrightarrow{\text{Simplex}}\; \text{SM}
\;\xrightarrow{\text{Block}}\; \text{Universe}
}
\end{equation}

\begin{center}
\begin{tabular}{rlll}
\hline
Step & From & To & How \\
\hline
0 & Nothing & Points with relations & (minimal assumption) \\
1 & Relations & $\mathbb{C}$ & Frobenius + phase + det \\
2 & $\mathbb{C}^N$ & $d = 5$ & Atomic uniqueness + swap \\
3 & $\mathbb{C}^5$ & SU(3)$\times$SU(2)$\times$U(1), $c\!=\!2$ & Chirality + density \\
4 & Hinges & $1 + 12 + 12 = 25$ & Binet--Cauchy + $c$-weight \\
5 & Channels & $\alpha_3, \alpha_2, \alpha_1$ & Basel sum + Gram rank \\
6 & $\alpha_i + v_H$ & 26 SM parameters & Simplex geometry \\
7 & $\hbar_h = 0$ (vacuum) & Cosmology (9 observables) & Block universe \\
8 & $\text{Tr}(G) = N$ & Ghost Sum Rule ($\Sigma\delta = 0$) & Completeness \\
\hline
& \multicolumn{2}{l}{\textbf{Free parameters:}} & \textbf{0} \\
& \multicolumn{2}{l}{\textbf{Observational inputs:}} & \textbf{0} \\
& \multicolumn{2}{l}{\textbf{Predictions:}} & \textbf{35+} \\
& \multicolumn{2}{l}{\textbf{Error structure:}} & \textbf{predicted (sum rule)} \\
\hline
\end{tabular}
\end{center}

\bigskip

\begin{center}
\emph{The universe is not a process. It is a matrix.}

\emph{Time is the direction we read it.}

\emph{Physics is the structure we find.}
\end{center}
