%=====================================================================
\section{QCD Phenomenology from $\mathbb{C}^5$}
\label{app:qcd}
%=====================================================================

This appendix develops the non-perturbative aspects of the strong force from the spatial sector $\mathbb{C}^3 \subset \mathbb{C}^5$. The key results --- confinement, asymptotic freedom, the deconfinement transition, and the strongly-coupled quark-gluon plasma (sQGP) --- all follow from the rank structure of the spatial Gram sub-matrix.

\subsection{The moment map: $\mathbb{CP}^4 \to \Delta^4$}

Each normalized state $\psi \in \mathbb{CP}^4$ maps to its probability vector:
\begin{equation}
\mu(\psi) = (|\psi_0|^2, \ldots, |\psi_4|^2) \in \Delta^4
\end{equation}
where $\Delta^4$ is the standard 4-simplex ($\sum_a |\psi_a|^2 = 1$). This is the moment map for the $\text{U}(1)^5$ torus action on $\mathbb{CP}^4$. The $(3,2)$ split gives:
\begin{equation}
\mu_S = \sum_{a=0}^{2}|\psi_a|^2 \quad (\text{spatial weight}), \qquad \mu_T = \sum_{a=3}^{4}|\psi_a|^2 \quad (\text{temporal weight}), \qquad \mu_S + \mu_T = 1.
\end{equation}

\subsection{Confinement as $\mathbb{C}^3$ saturation}

The spatial overlap between vertices $i$ and $j$:
\begin{equation}
G^S_{ij} = \sum_{a=0}^{2} \psi^*_{i,a}\,\psi_{j,a}
\end{equation}
has $|G^S_{ij}| \leq \mu_S(i)^{1/2}\,\mu_S(j)^{1/2}$ by Cauchy--Schwarz. The spatial Gram matrix $G^S$ has rank at most $n_S = 3$.

\textbf{Confinement criterion:} When 3 vertices $\{a, b, c\}$ span all of $\mathbb{C}^3$ (i.e., $\text{rank}(G^S_{abc}) = 3$), the spatial sector is \emph{saturated}. No additional vertex can be spatially independent. Any fourth vertex must be a linear combination of $\{a, b, c\}$ in the spatial sector --- it is ``confined'' to the color space already defined.

This is the lattice analogue of color confinement: quarks cannot exist as free particles because the $\mathbb{C}^3$ sector is always locally saturated. The confinement scale is not a free parameter but a geometric property of $n_S = 3$.

\subsection{Asymptotic freedom as $\mathbb{C}^5$ dilution}

At high energies (short distances), the full $\mathbb{C}^5$ becomes relevant. The effective coupling strength is:
\begin{equation}
\alpha_\text{eff}(Q) \propto \frac{n_S}{d} = \frac{3}{5}
\end{equation}
at the GUT scale, where all 5 dimensions participate equally. As energy decreases, the $(3,2)$ split freezes and the strong force sees only $\mathbb{C}^3$:
\begin{equation}
\alpha_\text{eff}(Q) \propto \frac{n_S}{n_S} = 1 \qquad (\text{low energy: confined}).
\end{equation}
The coupling grows from $3/5$ to $1$ as energy decreases --- this is asymptotic freedom.

In DRLT, asymptotic freedom is equivalent to Pachner coarse-graining: at lower resolution (fewer effective vertices), the spatial sector occupies a larger fraction of the available Hilbert space, increasing the effective coupling. The standard QCD $\beta$-function $b_3 = -7$ is the derivative of this geometric transition.

\subsection{Deconfinement: $\mathbb{C}^3 \to \mathbb{C}^5$}

At temperature $T > T_c$, thermal fluctuations overwhelm the spatial rank constraint. The $\mathbb{C}^3$ boundary dissolves: quarks and gluons access the full $\mathbb{C}^5$ simplex, including the temporal sector $\mathbb{C}^2$. All $d^2 = 25$ Binet--Cauchy channels (Chapter~\ref{ch:couplings}) activate simultaneously.

The order parameter is the Polyakov loop analogue:
\begin{equation}
L = \frac{1}{N}\sum_i \mu_S(i)
\end{equation}
which transitions from $\langle L\rangle \approx n_S/d = 3/5$ (confined) to $\langle L\rangle \to 1$ (deconfined, spatial sector dominates).

\textbf{Key constraint:} even in the deconfined phase, $\text{rank}(G) \leq 5$ is absolute. Particles cannot achieve infinitely many independent degrees of freedom. They \emph{must} share hinges with neighbors. This prevents ideal-gas behavior and forces collective flow.

\subsection{Entropy density from the Holevo bound}

Each hinge carries 1 bit of classical information (Holevo bound, Chapter~\ref{ch:hbar}, Theorem 4.1). In the sQGP, all $d^2 = 25$ channels are active, so the hinge is at maximum local information capacity. The entropy per unit hinge area in natural units:
\begin{equation}
s = \frac{S}{A} = \frac{1\;\text{bit}}{A_h} \qquad\Rightarrow\qquad S = A_h.
\end{equation}

\subsection{Shear viscosity from hinge absorption}

Momentum transfer between adjacent 4-simplices occurs \emph{only} through shared hinges --- the 2-dimensional triangular faces that connect them. The geometric absorption cross-section equals the hinge area $A_h$.

The maximum rate at which a gauge fluctuation can traverse a hinge is limited by two derived quantities:
\begin{itemize}
\item $c = 2$: the lattice propagation speed (Chapter~\ref{ch:geometry}, Eq.~\ref{eq:speed_of_light}), derived from $(n_S, n_T) = (3, 2)$.
\item $2\pi$: one complete $\text{U}(1)$ phase rotation --- the circumference of $S^1$, forced by $\mathbb{K} = \mathbb{C}$ (Chapter~\ref{ch:whyC}).
\end{itemize}

The total phase throughput of the hinge is $c \times 2\pi = 4\pi$. Shear viscosity is cross-section divided by phase throughput:
\begin{equation}
\eta = \frac{A_h}{c \times 2\pi} = \frac{A_h}{4\pi}.
\end{equation}
\textbf{Physical meaning:} the hinge is a geometric ``bottleneck'' with area $A_h$, through which gauge information passes at rate $c \times 2\pi$. Viscosity measures how effectively this bottleneck transmits momentum.

\subsection{The KSS bound}

Dividing $\eta$ by $s$, the hinge area cancels:

\begin{equation}
\boxed{\frac{\eta}{s} = \frac{A_h/(c \times 2\pi)}{A_h} = \frac{1}{c \times 2\pi} = \frac{1}{4\pi} \approx 0.0796.}
\end{equation}

This is the Kovtun--Son--Starinets (KSS) bound, derived here with:
\begin{itemize}
\item Zero free parameters.
\item No AdS/CFT correspondence.
\item No extra spatial dimension ($\mathbb{C}^5$ is the \emph{internal} space, not a 5th spatial direction).
\item $c = 2$: derived from vertex counting (Chapter~\ref{ch:geometry}).
\item $2\pi$: from $\text{U}(1) = S^1$ topology (Chapter~\ref{ch:whyC}).
\end{itemize}

\subsection{Why sQGP is a nearly perfect fluid}

\begin{theorem}
In DRLT, any deconfined state with all $d^2$ channels active has $\eta/s = 1/(4\pi)$.
\end{theorem}

\begin{proof}
In the deconfined phase, $\text{rank}(G) \leq 5$ remains absolute. All momentum transfer between simplices must pass through shared hinges with area $A_h$. The entropy saturates at 1 bit per hinge (Holevo). The viscosity is $A_h/(c \times 2\pi)$ (phase throughput). The ratio $\eta/s = 1/(c \times 2\pi) = 1/(4\pi)$ is independent of $A_h$, the temperature, and the number of vertices.
\end{proof}

\textbf{Physical explanation.} If $\text{rank}(G) \to \infty$ were possible, particles could bypass hinges entirely: $\eta \to \infty$ (ideal gas, not a fluid). But $\text{rank}(G) \leq 5$ is absolute: all momentum transfer must funnel through the $A_h$ bottleneck. This forces area-dependent collective flow. The sQGP is a ``nearly perfect fluid'' not because interactions are infinitely strong, but because the geometry is finitely constrained:

\begin{quote}
\emph{Freed from the $\mathbb{C}^3$ jail, but still inside the $\mathbb{C}^5$ prison.}
\end{quote}

\subsection{Comparison with AdS/CFT}

\begin{center}
\begin{tabular}{lll}
\hline
& AdS/CFT & DRLT \\
\hline
Method & 5D black hole dual & $\mathbb{C}^5$ simplex geometry \\
Extra dimension & Spatial (AdS$_5$) & Internal ($\mathbb{C}^5$) \\
Origin of $4\pi$ & Black hole horizon area & $c \times 2\pi$ (speed $\times$ U(1) period) \\
Free parameters & $\lambda \to \infty$ limit needed & 0 \\
Applies to & $\mathcal{N}{=}4$ SYM (not QCD) & Any deconfined $\mathbb{C}^5$ state \\
Result & $\eta/s \geq 1/(4\pi)$ (bound) & $\eta/s = 1/(4\pi)$ (exact value) \\
\hline
\end{tabular}
\end{center}

AdS/CFT gives an \emph{inequality}: $\eta/s \geq 1/(4\pi)$ in the infinite coupling limit of $\mathcal{N} = 4$ super Yang--Mills, a theory that is not QCD. DRLT gives the \emph{exact value} for any deconfined $\mathbb{C}^5$ configuration, with zero free parameters and no supersymmetry.

\subsection{Why AdS/CFT works: the 5 is the same 5}

The deepest question is not \emph{what} AdS/CFT computes, but \emph{why} it works at all. DRLT provides the answer: the 5 in AdS$_5$ is not an arbitrary choice --- it is $d = 5$, the dimension of the internal space $\mathbb{C}^5$.

\subsubsection{The dictionary}

\begin{center}
\begin{tabular}{lcl}
\hline
AdS/CFT & & DRLT \\
\hline
AdS$_5$ (5D spacetime) & $=$ & $\mathbb{C}^5$ (internal space, $d = 5$) \\
Black hole horizon & $=$ & $\det(G_h) \to 0^+$ boundary \\
Hawking temperature & $=$ & $\hbar_\text{eff} \to 0^+$ (Chapter~\ref{ch:hbar}) \\
Strong coupling ($\lambda \to \infty$) & $=$ & All $d^2 = 25$ channels active \\
Horizon area & $=$ & Hinge area $A_h$ \\
Bekenstein--Hawking entropy & $=$ & 1 bit per hinge (Holevo) \\
\hline
\end{tabular}
\end{center}

\subsubsection{The black hole horizon IS the sQGP}

In DRLT, a black hole horizon is a surface where $\det(G_h) \to 0^+$: vertices are nearly aligned ($\psi_i \approx e^{i\theta}\psi_j$), $\hbar_\text{eff} \to 0$, and the region transitions from quantum to classical (Chapter~\ref{ch:hbar}). The sQGP is a state where the full $\mathbb{C}^5$ is active and energy density is extreme, driving $\det(G_h)$ toward uniformity and then toward zero.

These are the \emph{same geometric state}:

\begin{center}
\begin{tabular}{lccc}
\hline
& Region & $\det(G_h)$ & $\hbar_\text{eff}$ \\
\hline
\multicolumn{4}{l}{\emph{Black hole:}} \\
\quad Exterior & quantum & $> 0$ & $> 0$ \\
\quad Horizon & transition & $\to 0^+$ & $\to 0^+$ \\
\quad Interior & vacuum ($\psi$ aligned) & $= 0$ & $= 0$ \\
\hline
\multicolumn{4}{l}{\emph{Heavy-ion collision:}} \\
\quad $T < T_c$ & confined (hadrons) & large, fluctuating & $> 0$ \\
\quad $T = T_c$ & deconfined (sQGP) & uniform, $\to 0^+$ & $\to 0^+$ \\
\quad $T \to \infty$ & compressed $\to$ aligned & $\to 0$ & $\to 0$ \\
\hline
\end{tabular}
\end{center}

The physical environments are different --- one is gravitational collapse, the other is nuclear collision --- but the Gram matrix structure is identical: \textbf{$\mathbb{C}^5$ fully active, $\det(G_h) \to 0^+$, all 25 channels open}.

\subsubsection{Why AdS/CFT had to use 5 dimensions}

AdS/CFT did not know \emph{why} 5 dimensions work. The standard justification is mathematical: the conformal group $\text{SO}(4,2)$ of 4D CFT is the isometry group of AdS$_5$. In DRLT, the deeper reason is:

\begin{enumerate}
\item The internal space is $\mathbb{C}^d$ with $d = 5$ (derived, Chapter~\ref{ch:whyd5}).
\item A black hole horizon in $\mathbb{C}^5$ has $\det(G_h) \to 0^+$, which is geometrically identical to the sQGP state.
\item Computing $\eta/s$ on this surface gives $1/(c \times 2\pi) = 1/(4\pi)$ by hinge area cancellation.
\item AdS/CFT reproduces this because it is (unknowingly) computing on the same $d = 5$ geometry.
\end{enumerate}

The AdS/CFT correspondence is not a ``duality'' between two different theories. It is a \emph{shadow} of the fact that the internal space is $\mathbb{C}^5$, projected into the language of 5-dimensional general relativity. It works because the 5 is right. It requires $\lambda \to \infty$ because only in that limit does the shadow faithfully reproduce the underlying $\mathbb{C}^5$ geometry.

\subsection{The QCD phase diagram from $\mathbb{C}^3$}

\begin{center}
\begin{tabular}{lccl}
\hline
Regime & $G^S$ rank & $\eta/s$ & Physics \\
\hline
Confined (hadrons) & 3 (locally saturated) & large & Color singlets only \\
Crossover ($T \sim T_c$) & 3 (globally spreading) & decreasing & Deconfinement \\
sQGP ($T > T_c$) & 3 (global, all 25 ch.) & $1/(4\pi)$ & Nearly perfect fluid \\
Asymptotically free & 3 (diluted in $\mathbb{C}^5$) & increasing & Weakly coupled gas \\
\hline
\end{tabular}
\end{center}

The entire QCD phase structure --- confinement, deconfinement, the sQGP as a nearly perfect fluid, and asymptotic freedom --- follows from one integer ($n_S = 3$) and two derived constants ($c = 2$, $2\pi$).
