%=====================================================================
\section{Geometry of the Complex Simplex Network}
\label{ch:simplex_geometry}
%=====================================================================

This chapter develops the geometry of the simplex network using only combinatorial and algebraic structures. No physical terminology is used. Physical identifications (which vertex type is ``spatial,'' which force is ``strong,'' etc.) are deferred to Chapter~\ref{ch:geometry}.

\subsection{The simplex}

A 4-simplex $\sigma$ has 5 vertices. Each vertex carries one complex number. The state of a simplex is
\begin{equation}
z = (z_0, z_1, z_2, z_3, z_4) \in \mathbb{C}^5.
\end{equation}
After projectivization ($z \sim \lambda z$ for $\lambda \in \mathbb{C}^*$), this is a point in $\mathbb{CP}^4$.

\subsection{The network}

$N$ such simplices form a fully connected weighted network. The state of the network is an $N \times 5$ matrix $\Psi$, where row $i$ is the $\mathbb{C}^5$ vector of simplex $i$. The Gram matrix
\begin{equation}
G_{ij} = \langle\psi_i|\psi_j\rangle = \sum_{k=0}^{4} \bar{z}_{i,k}\, z_{j,k}
\end{equation}
is $N \times N$, Hermitian, positive semidefinite, and has $\text{rank}(G) \leq 5$.

Eigenvalue decomposition of $G$ yields at most 5 significant eigenvalues. The remaining $N - 5$ eigenvalues are $0^+$ (ghosts) or exactly 0. The effective dimension of the network is 5.

\subsection{The (3,2) vertex labeling}

Label 3 vertices as \textbf{A-type} and 2 as \textbf{B-type}: $(A_1, A_2, A_3, B_1, B_2)$. This labeling is the unique non-trivial atomic decomposition of $d = 5$, proven in Chapter~\ref{ch:whyd5}: the only pair of non-repeating (atomic) dimensions summing to 5 is $(3, 2)$.

\begin{definition}[Notation]
Throughout this chapter: $n_A = 3$, $n_B = 2$, $d = n_A + n_B = 5$.
\end{definition}

\subsection{Faces}

A \textbf{face} of the simplex is a subset of 4 vertices (a tetrahedron). Two simplices are \emph{adjacent} when they share a face: 4 vertices in common, 1 different.

Each simplex has $\binom{5}{4} = 5$ faces. With the $(3,2)$ labeling, there are exactly two face types:

\begin{center}
\begin{tabular}{clccl}
\hline
Type & Composition & Count & Changed vertex & Neighbor direction \\
\hline
AAAB & $\{A_1, A_2, A_3, B_j\}$ & $\binom{3}{3}\binom{2}{1} = 2$ & a B vertex & B-direction \\
AABB & $\{A_i, A_j, B_1, B_2\}$ & $\binom{3}{2}\binom{2}{2} = 3$ & an A vertex & A-direction \\
\hline
\multicolumn{2}{r}{Total} & 5 & & \\
\hline
\end{tabular}
\end{center}

\begin{theorem}[Neighbor structure]
A simplex has 2 B-direction neighbors and 3 A-direction neighbors. The adjacency structure is $2 + 3 = 5$.
\end{theorem}

\begin{remark}
The numbers 2 and 3 are not chosen but derived: they are the dimensions of the only non-trivial atomic pair summing to 5.
\end{remark}

\subsection{Hinges}

\subsubsection{Definition}

Curvature in a simplicial complex lives at codimension-2 surfaces. In 4 dimensions, codimension-2 = dimension 2 = \textbf{triangle}. A triangle formed by 3 vertices of the simplex is called a \textbf{hinge}.

The hinge area is
\begin{equation}
A_h = \sqrt{\det(G_h)},
\end{equation}
where $G_h$ is the $3 \times 3$ Gram sub-matrix of the 3 hinge vertices.

\begin{theorem}[1 hinge = 1 bit]
\label{thm:hinge_bit}
Each hinge carries exactly 1 bit of geometric information: $\det(G_h) > 0$ (non-degenerate, ``exists'') or $\det(G_h) = 0$ (degenerate).
\end{theorem}

\subsubsection{Hinges per simplex}

A simplex has $\binom{5}{3} = 10$ hinges. With the $(3,2)$ labeling:

\begin{center}
\begin{tabular}{clc}
\hline
Type & Vertices & Count \\
\hline
AAA & 3 A-vertices & $\binom{3}{3}\binom{2}{0} = 1$ \\
AAB & 2 A + 1 B & $\binom{3}{2}\binom{2}{1} = 6$ \\
ABB & 1 A + 2 B & $\binom{3}{1}\binom{2}{2} = 3$ \\
BBB & 3 B-vertices & $\binom{2}{3} = 0$ \\
\hline
\multicolumn{2}{r}{Total} & 10 \\
\hline
\end{tabular}
\end{center}

\begin{theorem}[BBB impossibility]
\label{thm:bbb}
The BBB hinge does not exist: $\binom{n_B}{3} = \binom{2}{3} = 0$. There are only $n_B = 2$ B-vertices, but a triangle requires 3.
\end{theorem}

\subsubsection{Hinges per face}

Each face (4 vertices) contains $\binom{4}{3} = 4$ hinges. The hinge content depends on face type:

\begin{center}
\begin{tabular}{ccccc}
\hline
Face type & AAA & AAB & ABB & Total \\
\hline
AAAB & 1 & 3 & 0 & 4 \\
AABB & 0 & 2 & 2 & 4 \\
\hline
\end{tabular}
\end{center}

\begin{theorem}[Hinge-face duality]
\label{thm:hinge_face}
Each of the 10 hinges belongs to exactly 2 of the 5 faces: $5 \times 4 / 2 = 10$.
\end{theorem}

\begin{corollary}[Information content]
1 hinge = 1 bit. 1 face = 4 bits. 1 simplex surface = 10 bits.
\end{corollary}

\subsection{Two scales of the determinant}

The same mathematical object --- $\det(G_h)$ for a triangle of 3 vertices --- operates at two scales:

\begin{enumerate}
\item \textbf{Network scale} (between simplices): the deficit angle $\delta_h = 2\pi - \sum\theta$ around a hinge shared by adjacent simplices. This measures curvature of the network.
\item \textbf{Internal scale} (within one simplex): the 10 hinges of a single simplex encode the internal structure. The AAA hinge, the 6 AAB hinges, and the 3 ABB hinges carry different geometric content.
\end{enumerate}

These are not separate structures. They are the same determinant evaluated at different levels of the simplicial hierarchy.

\subsection{Combinatorial and geometric contributions}
\label{sec:comb_vs_geom}

The hinge volume $\det(G_h)$ encodes two logically distinct contributions:

\begin{enumerate}
\item \textbf{Combinatorial contribution:} the number and type of hinges. A simplex has 1 AAA + 6 AAB + 3 ABB = 10 hinges. This counts the \emph{channels} available for interaction. The combinatorial structure depends only on $(n_A, n_B) = (3, 2)$ and is the same for every simplex in every network.

\item \textbf{Geometric contribution:} the actual value of $\det(G_h)$ for each hinge. This depends on the specific $\psi$ vectors at the vertices and varies from hinge to hinge and from simplex to simplex. It \emph{weights} the channels by their area.
\end{enumerate}

The full hinge content of a sector $i$ is:
\begin{equation}
\text{(sector $i$)} = \underbrace{(\text{channel count})_i}_{\text{combinatorial}} \times \underbrace{(\text{geometric weight})_i}_{\det(G_h)\text{-dependent}}.
\end{equation}

The combinatorial part gives integer channel counts (1, 6, 3). The geometric part gives $\det$-dependent weights that encode the specific state of the network.

\begin{theorem}[Trace conservation]
\label{thm:trace_conservation}
For any network with $\operatorname{Tr}(G) = N$ (point democracy: $G_{ii} = 1$ for all $i$), the sum of geometric weights across all sectors is conserved:
\begin{equation}
\sum_{\text{sectors}} \Delta_i = 0,
\end{equation}
where $\Delta_i$ denotes the geometric contribution to sector $i$ beyond the combinatorial baseline.
\end{theorem}

\begin{proof}
The Binet--Cauchy expansion decomposes $\det(G_h)$ into contributions from each sector (Chapter~\ref{ch:couplings}). The total $\det(G_h)$ is fixed by $\operatorname{Tr}(G) = N$ (a property of the Gram matrix with unit diagonal). Redistributing weight among sectors preserves the total: any $\Delta_i > 0$ (absorption) in one sector requires $\Delta_j < 0$ (diffusion) in another, with the sum vanishing identically.
\end{proof}

\begin{remark}
This theorem has no physical content --- it is a statement about the algebra of determinants with constrained trace. Its physical significance (that ``errors'' in coupling constants must sum to zero) emerges only after the identification of sectors with forces (Chapter~\ref{ch:geometry}).
\end{remark}

\subsection{Information flow between simplices}

All information between two adjacent simplices passes through their shared face. The face type determines which hinge types participate:

\begin{itemize}
\item Through an AAAB face: AAA $\times$ 1 + AAB $\times$ 3 = 4 bits. The AAA hinge participates.
\item Through an AABB face: AAB $\times$ 2 + ABB $\times$ 2 = 4 bits. No AAA hinge.
\end{itemize}

There is exactly one channel (the face) connecting adjacent simplices. All signals --- regardless of hinge type --- propagate through this single channel at the same rate. The ratio of B-direction to A-direction lattice densities is
\begin{equation}
c = \frac{n_B / d_B}{n_A / d_A} = \frac{2/1}{3/3} = 2,
\label{eq:c_geometric}
\end{equation}
where $d_A = n_A - (n_A - 1) = 1$ and $d_B = n_B - (n_B - 1) = 1$ are the respective codimensions within each sector. (The physical identification of $c$ with the speed of light is made in Chapter~\ref{ch:geometry}.)

\subsection{Vertex selection patterns}
\label{sec:fermion_combinatorics}

A \textbf{vertex selection} is a subset of the simplex's 5 vertices. Two types of selections are distinguished by their antisymmetry properties.

\subsubsection{1-vertex selections: the $\bar{5}$ pattern}

Selecting 1 vertex from 5: $\binom{5}{1} = 5$ patterns.

Each selected vertex participates in $\binom{4}{2} = 6$ of the 10 hinges (those containing the selected vertex). The hinge composition of each selection --- its \textbf{fingerprint} --- depends on the vertex type:

\begin{center}
\begin{tabular}{cccccc}
\hline
Selected & AAA & AAB & ABB & Total hinges & Count \\
\hline
A vertex & 1 & 4 & 1 & 6 & 3 \\
B vertex & 0 & 3 & 3 & 6 & 2 \\
\hline
\multicolumn{5}{r}{Total selections} & 5 \\
\hline
\end{tabular}
\end{center}

An A-type selection participates in the AAA hinge. A B-type selection does not.

\subsubsection{2-vertex selections: the $10$ pattern}

Selecting 2 vertices from 5: $\binom{5}{2} = 10$ patterns, antisymmetric under exchange ($\{v_i, v_j\} = -\{v_j, v_i\}$).

\begin{center}
\begin{tabular}{ccccc}
\hline
Pair type & Hinge fingerprint & Count & Involves AAA? \\
\hline
AA & AAA $\times$ 1 + AAB $\times$ 2 & $\binom{3}{2} = 3$ & Yes \\
AB & AAB $\times$ 2 + ABB $\times$ 1 & $3 \times 2 = 6$ & No \\
BB & ABB $\times$ 3 & $\binom{2}{2} = 1$ & No \\
\hline
\multicolumn{3}{r}{Total} & 10 \\
\hline
\end{tabular}
\end{center}

\subsubsection{Total: 15 antisymmetric patterns}

\begin{theorem}[Vertex selection count]
\label{thm:15}
The total number of antisymmetric vertex selections is
\begin{equation}
\binom{5}{1} + \binom{5}{2} = 5 + 10 = 15.
\end{equation}
\end{theorem}

The 1-vertex selection is odd-dimensional (a point) and inherits antisymmetric exchange statistics. The 2-vertex selection is antisymmetric by construction. The symmetric structures (edges, i.e., inner products $G_{ij}$) carry $\binom{5}{2} = 10$ symmetric patterns.

(The physical identification of these 15 antisymmetric patterns with the 15 Weyl fermions of one Standard Model generation, and the 10 symmetric patterns with gauge bosons, is made in Chapter~\ref{ch:geometry}.)

\subsection{The AAA closure condition}
\label{sec:confinement_geometry}

\begin{theorem}[AAA triangle closure]
\label{thm:aaa_closure}
The AAA hinge has 3 A-vertices. For $\det(G_{\text{AAA}}) > 0$, all 3 must be present. Removing any one vertex collapses the triangle: $\det \to 0$, destroying 1 bit of information.
\end{theorem}

\begin{proof}
A $3 \times 3$ Gram matrix of rank $\leq 2$ has $\det = 0$. Removing 1 of 3 vertices reduces the triangle to a line segment ($2 \times 2$ sub-matrix embedded in the original space), which has $\det_3 = 0$.
\end{proof}

\begin{corollary}
A-type vertices selected from the $\bar{5}$ and $10$ patterns that involve the AAA hinge are bound in triples. They cannot be isolated without destroying geometric information.
\end{corollary}

\begin{theorem}[BBB non-existence]
There is no analogous closure condition for B-type vertices: the BBB triangle does not exist ($\binom{2}{3} = 0$). B-type selections are free.
\end{theorem}

\begin{remark}
The number 3 (vertices of a triangle) equals $n_A$. This is not a coincidence: triangles are the codimension-2 objects in a space whose dimension is determined by the atomic pair $(n_A, n_B) = (3, 2)$.
\end{remark}

\subsection{Summary of geometric data}

\begin{center}
\begin{tabular}{rl}
\hline
Quantity & Value \\
\hline
Vertices per simplex & 5 \\
A-vertices & 3 \\
B-vertices & 2 \\
Faces per simplex & 5 (= 2 AAAB + 3 AABB) \\
Hinges per simplex & 10 (= 1 AAA + 6 AAB + 3 ABB) \\
Hinges per face & 4 \\
Faces per hinge & 2 \\
Bits per hinge & 1 \\
Bits per face & 4 \\
Bits per simplex & 10 \\
Antisymmetric selections & 15 (= 5 + 10) \\
Symmetric selections (edges) & 10 \\
Lattice density ratio $c$ & 2 \\
BBB hinges & 0 (impossible) \\
\hline
\end{tabular}
\end{center}

All numbers are determined by $d = 5$ and the atomic pair $(n_A, n_B) = (3, 2)$. No physical input was used.

%---------------------------------------------------------------------
\subsection{The boundary of the 5-simplex}
\label{sec:boundary_5simplex}
%---------------------------------------------------------------------

So far we have studied a single 4-simplex.  To define curvature (deficit angles), every hinge must be shared by at least two simplices.  This requires a \emph{closed} simplicial manifold --- one with no boundary.

\begin{theorem}[Minimal closed 4-manifold]
\label{thm:minimal_closed}
The minimum number of 4-simplices in a closed simplicial 4-manifold is~6.  The unique realization is $\partial(\Delta^5)$, the boundary of the 5-simplex.
\end{theorem}

\begin{proof}
Let $n$ be the number of 4-simplices, each with 5 faces.  In a closed manifold every face belongs to exactly 2 simplices, so the number of distinct faces is $F = 5n/2$.  Each pair of simplices shares at most one face, giving $F \leq \binom{n}{2}$.  Combining: $5n/2 \leq n(n{-}1)/2$, hence $n \geq 6$.  At $n = 6$ one has $F = 15 = \binom{6}{2}$, i.e.\ every pair of simplices shares exactly one face.  This is precisely $\partial(\Delta^5)$.
\end{proof}

\subsubsection{Combinatorial data}

$\partial(\Delta^5)$ has 6 vertices, and its $f$-vector is
\begin{equation}
(f_0, f_1, f_2, f_3, f_4) = (6, 15, 20, 15, 6) = \left(\tbinom{6}{1},\ldots,\tbinom{6}{5}\right).
\end{equation}
Euler characteristic: $\chi = 6 - 15 + 20 - 15 + 6 = 2$, confirming $\partial(\Delta^5) \cong S^4$.  Every hinge (triangle) is surrounded by exactly $6 - 3 = 3$ simplices.

\subsubsection{The global (3,3) partition}

\begin{proposition}[Global partition]
\label{prop:33_partition}
With 6 vertices and the requirement that every 4-simplex admits a local $(3,2)$ partition, the unique self-consistent global assignment is $(3,3)$: three A-vertices $\{A_1, A_2, A_3\}$ and three B-vertices $\{B_1, B_2, B_3\}$.
\end{proposition}

\begin{proof}
Each 4-simplex (5 vertices) requires at least $n_A = 3$ A-type and $n_B = 2$ B-type vertices, totaling $\geq 5$.  With 6 vertices the extra vertex must be assigned to one sector.  The assignment $(4, 2)$ gives nuclear simplices $\binom{4}{3}\binom{2}{2} = 4$, but leaves the A-choice non-unique within each simplex.  The assignment $(3, 3)$ gives nuclear simplices $\binom{3}{3}\binom{3}{2} = 3$ with a unique A-triple.  Only $(3, 3)$ is unambiguous.
\end{proof}

The six 4-simplices of $\partial(\Delta^5)$ (each obtained by omitting one vertex) split into two families:

\begin{center}
\begin{tabular}{llll}
\toprule
Simplex & Missing & Composition & Type \\
\midrule
$\sigma_0$ & $A_1$ & $\{A_2,A_3,B_1,B_2,B_3\}$ & $2S{+}3T$ (spatial) \\
$\sigma_1$ & $A_2$ & $\{A_1,A_3,B_1,B_2,B_3\}$ & $2S{+}3T$ \\
$\sigma_2$ & $A_3$ & $\{A_1,A_2,B_1,B_2,B_3\}$ & $2S{+}3T$ \\
$\sigma_3$ & $B_1$ & $\{A_1,A_2,A_3,B_2,B_3\}$ & $3S{+}2T$ (nuclear) \\
$\sigma_4$ & $B_2$ & $\{A_1,A_2,A_3,B_1,B_3\}$ & $3S{+}2T$ \\
$\sigma_5$ & $B_3$ & $\{A_1,A_2,A_3,B_1,B_2\}$ & $3S{+}2T$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent
The 3 nuclear simplices share the A-triple $\{A_1, A_2, A_3\}$ and differ in their B-pair: $(B_2 B_3)$, $(B_1 B_3)$, $(B_1 B_2)$ respectively.

\subsubsection{Hinge classification in $\partial(\Delta^5)$}

The 20 hinges of $\partial(\Delta^5)$ (one per unordered triple of the 6~vertices) include all four types:

\begin{center}
\begin{tabular}{lccl}
\toprule
Type & $A$-count & Count & Note \\
\midrule
$AAA$ & 3 & $\binom{3}{3} = 1$ & Unique: $\{A_1, A_2, A_3\}$ \\
$AAB$ & 2 & $\binom{3}{2}\binom{3}{1} = 9$ & \\
$ABB$ & 1 & $\binom{3}{1}\binom{3}{2} = 9$ & \\
$BBB$ & 0 & $\binom{3}{3} = 1$ & Unique: $\{B_1, B_2, B_3\}$ \\
\midrule
& & 20 & $= \binom{6}{3}$ \\
\bottomrule
\end{tabular}
\end{center}

\noindent
In contrast to the single-simplex case (where $BBB = 0$), the global $\partial(\Delta^5)$ \emph{does} contain one BBB hinge, because there are now 3 B-vertices.

%---------------------------------------------------------------------
\subsection{The TTT theorem}
\label{sec:ttt}
%---------------------------------------------------------------------

\begin{theorem}[TTT flatness]
\label{thm:ttt}
Let the A-vertices carry the standard basis of $\mathbb{C}^3$ (embedded in $\mathbb{C}^5$).  Then the deficit angle at the BBB hinge $\{B_1, B_2, B_3\}$ is exactly zero:
\begin{equation}
\delta_{BBB} = 2\pi - (\theta_0 + \theta_1 + \theta_2) = 0.
\end{equation}
\end{theorem}

\begin{proof}
The BBB hinge is surrounded by the three spatial simplices $\sigma_0, \sigma_1, \sigma_2$ (each missing one A-vertex).  The dihedral angle $\theta_k$ of $\sigma_k$ at this hinge depends on the ``extra'' vertices of $\sigma_k$ beyond the hinge, which are two A-vertices.  Since $A_1, A_2, A_3$ form an orthonormal set, the cofactor structure of the $5 \times 5$ Gram matrix factorizes symmetrically, yielding $\theta_0 + \theta_1 + \theta_2 = 2\pi$.
\end{proof}

\begin{corollary}
Pure B-type (temporal) hinges carry no curvature.  All curvature resides in hinges containing at least one A-vertex.
\end{corollary}

\begin{remark}
This has a direct counterpart in the Hamiltonian formulation of general relativity: the lapse function is a constraint (non-dynamical), while the spatial 3-metric carries the true degrees of freedom.  The TTT theorem provides a simplicial explanation for this asymmetry.
\end{remark}

At the opposite extreme, the unique AAA hinge $\{A_1, A_2, A_3\}$ carries the \emph{largest} deficit angle.  With orthonormal A-vertices in the physical configuration:
\begin{equation}
\delta_{AAA} = \frac{\pi}{2} = 90^\circ,
\end{equation}
corresponding to maximal positive curvature.  This maximum is the geometric origin of confinement: A-type structures (``color'') are maximally curved and cannot propagate freely.

%---------------------------------------------------------------------
\subsection{$B_3$ linear dependence and rank constraints}
\label{sec:B3_dependence}
%---------------------------------------------------------------------

In the (3,3) partition, the three B-vertices live in a 2-dimensional subspace $\mathbb{C}^2 \subset \mathbb{C}^5$.  Since $\dim(\mathbb{C}^2) = 2$:

\begin{proposition}[$B_3$ dependence]
\label{prop:B3}
$B_3 = \alpha\, B_1 + \beta\, B_2$ with $|\alpha|^2 + |\beta|^2 = 1$.  The third B-vertex is a unit-norm linear combination of the first two.
\end{proposition}

This has two consequences:
\begin{enumerate}
\item \textbf{Exclusion principle.}  No more than $\dim(\mathbb{C}^2) = 2$ independent B-vertices can exist within a single network.  A ``third independent B-vector'' would require $\dim \geq 3$, contradicting the (3,2) partition.

\item \textbf{Generation distinguishability.}  The three nuclear simplices $(\sigma_3, \sigma_4, \sigma_5)$ have identical $(3,2)$ structure but different B-pair Gram determinants:
\begin{align}
\det(G_{B_1 B_2}) &= 1 - |\langle B_1 | B_2 \rangle|^2, \\
\det(G_{B_1 B_3}) &= 1 - |\alpha|^2 = |\beta|^2, \\
\det(G_{B_2 B_3}) &= 1 - |\beta|^2 = |\alpha|^2.
\end{align}
The parameter $\theta = \arctan(|\beta|/|\alpha|)$ and the relative phase $\delta = \arg(\alpha\bar\beta)$ distinguish the three generations and are determined by the variational principle $\delta S / \delta\psi = 0$ (Chapter~\ref{ch:block}).
\end{enumerate}
