%=====================================================================
\section{Geometry and Gauge from the Gram Matrix}
\label{ch:geometry}
%=====================================================================

Chapters~\ref{ch:whyC}--\ref{ch:whyd5} established $\psi_i \in \mathbb{C}^5$ with $\|\psi_i\| = 1$. We now extract the full content of general relativity and the Standard Model gauge structure from the Gram matrix $G_{ij} = \langle\psi_i|\psi_j\rangle$.

\subsection{The Gram matrix and its properties}

The inner products between all $N$ points define the Gram matrix:
\begin{equation}
G_{ij} = \langle\psi_i|\psi_j\rangle = \sum_{a=0}^{4} \psi^*_{i,a}\,\psi_{j,a}.
\end{equation}

This $N \times N$ matrix has the following properties, all of which are \emph{theorems} (consequences of the inner product structure), not assumptions:

\begin{enumerate}[label=(\alph*)]
\item \textbf{Hermitian:} $G^\dagger = G$, since $G_{ji} = G^*_{ij}$.
\item \textbf{Positive semidefinite:} $\mathbf{c}^\dagger G\,\mathbf{c} = \|\Psi^\dagger\mathbf{c}\|^2 \geq 0$ for all $\mathbf{c} \in \mathbb{C}^N$.
\item \textbf{Rank $\leq 5$:} $G = \Psi\Psi^\dagger$ where $\Psi$ is $N \times 5$, so $\text{rank}(G) \leq 5$.
\item \textbf{Unit diagonal:} $G_{ii} = \|\psi_i\|^2 = 1$ (point democracy).
\item \textbf{Bounded:} $|G_{ij}| \leq 1$ (Cauchy--Schwarz).
\end{enumerate}

Each element $G_{ij}$ is a complex number carrying $2d = 10$ real parameters (a sum of $d = 5$ complex products $\psi^*_{i,a}\psi_{j,a}$). The number of independent components of a 4-dimensional symmetric metric tensor $g_{\mu\nu}$ is $d_\text{st}(d_\text{st}+1)/2 = 10$. This coincidence is exact: one edge of $G$ carries precisely the information content of one 4D metric.

\subsection{Polar decomposition: geometry and gauge}

Each off-diagonal element decomposes into magnitude and phase:
\begin{equation}
G_{ij} = |G_{ij}|\,e^{i\phi_{ij}}, \qquad |G_{ij}| \in [0, 1], \quad \phi_{ij} \in (-\pi, \pi].
\end{equation}

Define the geometric and gauge variables:
\begin{align}
W_{ij} &= \frac{|G_{ij}|^2}{d} \qquad\text{(real, symmetric: } W_{ij} = W_{ji}\text{)}, \label{eq:W}\\
\phi_{ij} &= \arg(G_{ij}) \qquad\text{(real, antisymmetric: } \phi_{ji} = -\phi_{ij}\text{)}. \label{eq:phi}
\end{align}

\textbf{$W$ encodes geometry} (Sec.~\ref{sec:metric}). \textbf{$\phi$ encodes gauge fields} (Sec.~\ref{sec:gauge_connection}).

This separation is the lattice version of the statement: ``gravity is the symmetric part of the connection; gauge forces are the antisymmetric part.''

\subsection{Metric and distance}
\label{sec:metric}

The geodesic distance on $\mathbb{CP}^4$ between points $i$ and $j$ is the Fubini--Study distance:
\begin{equation}
d^2_\text{FS}(i, j) = \arccos^2|G_{ij}|.
\end{equation}

For nearby points ($|G_{ij}| \approx 1$), this linearizes to a metric:
\begin{equation}
ds^2_{ij} = 1 - d\,W_{ij}.
\end{equation}

The interval is spacelike ($ds^2 > 0$) when $W_{ij} < 1/d$, and null ($ds^2 = 0$) when $|G_{ij}| = 1$ (identical points). No two distinct points can have $ds^2 \leq 0$ for generic $\psi$, which eliminates closed timelike curves.

\subsection{Simplicial structure}
\label{sec:simplex}

The rank constraint $\text{rank}(G) \leq 5$ means at most 5 points are linearly independent. Any set of 5 generic points in $\mathbb{CP}^4$ forms a \emph{4-simplex} (pentachoron). The geometry is naturally simplicial: Regge calculus emerges without being imposed.

A 4-simplex has:
\begin{equation}
\binom{5}{1} = 5 \text{ vertices}, \quad
\binom{5}{2} = 10 \text{ edges}, \quad
\binom{5}{3} = 10 \text{ hinges (triangles)}, \quad
\binom{5}{4} = 5 \text{ tetrahedra}, \quad
\binom{5}{5} = 1 \text{ 4-simplex}.
\end{equation}

\subsection{The simplex as data bus: 10 edges = 10 metric components}
\label{sec:data_bus}

The 4-simplex has $\binom{5}{2} = 10$ edges. A 4-dimensional symmetric metric tensor $g_{\mu\nu}$ has $d_\text{st}(d_\text{st}+1)/2 = 10$ independent components. This is not a coincidence: each edge of the simplex maps to one component of the metric tensor via the Regge correspondence.

Each edge carries \emph{two} data channels from the full inner product $G_{ij} = |G_{ij}|\,e^{i\phi_{ij}}$:
\begin{itemize}
\item $|G_{ij}|^2 \to W_{ij}$: the metric component (\textbf{gravity}),
\item $\phi_{ij} = \arg(G_{ij})$: the gauge connection (\textbf{forces}).
\end{itemize}

Under the causal $(n_A, n_B) = (3, 2)$ split, the 10 edges decompose as:

\begin{center}
\begin{tabular}{lccl}
\hline
Edge type & Count & Formula & Metric role \\
\hline
Temporal--Temporal & $\binom{2}{2} = 1$ & $W$ within time & $g_{00}$ (lapse) \\
Spatial--Spatial & $\binom{3}{2} = 3$ & $W$ within space & $g_{ij},\;i\neq j$ (spatial off-diagonal) \\
Spatial--Temporal & $2\times 3 = 6$ & $W$ across sectors & $g_{0i}$ (shift) + diagonal \\
\hline
\textbf{Total} & \textbf{10} & & $g_{\mu\nu}$ fully determined \\
\hline
\end{tabular}
\end{center}

The simplex is a minimal processor: 5 complex-number registers (vertices) connected by 10 pairwise comparisons (edges) whose magnitudes encode local geometry and whose phases encode gauge fields. The simplex interior carries no information --- it is the irreducible ``blind spot'' that defines the minimum resolution unit.

\subsection{Lorentz signature from unitarity}
\label{sec:Lorentz}

The metric $ds^2_{ij} = 1 - d\,W_{ij}$ is positive-definite (Riemannian). All distances are spacelike. Where does the Lorentzian signature $(-,+,+,+)$ come from?

\begin{theorem}[Lorentz signature is derived]
Unitarity of time evolution forces Lorentzian signature.
\end{theorem}

\begin{proof}
The argument has four steps.

\medskip
\noindent\textbf{Step 1: Unitarity selects a time direction.}
Time evolution between spatial slices $\Sigma_t$ and $\Sigma_{t+1}$ is unitary: $U^\dagger U = I$ (a consequence of inner-product preservation). This foliation breaks the directional symmetry of $G$: directions \emph{within} a slice are spatial; directions \emph{between} slices are temporal.

\medskip
\noindent\textbf{Step 2: The factor of $i$ is non-negotiable.}
The unitary propagator $U(\Delta t) = e^{-iH\Delta t}$ requires the imaginary unit. The alternatives fail: $e^{-H\Delta t}$ decays (not unitary), $e^{+H\Delta t}$ grows (not unitary). Only $e^{-iH\Delta t}$ oscillates, conserving probabilities.

\medskip
\noindent\textbf{Step 3: Wick rotation is forced.}
The path integral weight $e^{iS/\hbar}$ relates to the Euclidean weight $e^{-S_E/\hbar}$ via $\tau = it$. This Wick rotation transforms the Euclidean metric $ds^2_E = d\tau^2 + d\mathbf{x}^2$ into the Lorentzian metric $ds^2_L = -dt^2 + d\mathbf{x}^2$: the temporal component flips sign.

\medskip
\noindent\textbf{Step 4: Lattice metric.}
On the lattice, spacelike edges (within a slice) retain $ds^2_\text{space} = 1 - d\,W_{ij} > 0$, while timelike edges (between slices) acquire $ds^2_\text{time} = -(1 - d\,W_{ij}) < 0$. The combined signature is $(-,+,+,+)$.
\end{proof}

\begin{remark}
The light cone corresponds to $ds^2 = 0$, i.e., $|G_{ij}| = 1$: the boundary where two points become indistinguishable. Lorentz invariance $\text{SO}(1,3)$ is the symmetry group preserving this derived signature --- it is a theorem, not a postulate. Special and general relativity follow as the flat and curved limits, respectively.
\end{remark}

\subsection{Hinge area and curvature}

A hinge (triangle) with vertices $\{i, j, k\}$ has a $3\times 3$ Gram sub-matrix:
\begin{equation}
G_h = \begin{pmatrix} 1 & G_{ij} & G_{ik} \\ G_{ji} & 1 & G_{jk} \\ G_{ki} & G_{kj} & 1 \end{pmatrix}.
\end{equation}

Its area in $\mathbb{CP}^4$ is:
\begin{equation}
A_h = \sqrt{\det(G_h)}.
\end{equation}

Expanding the determinant:
\begin{equation}
\det(G_h) = 1 - |G_{ij}|^2 - |G_{jk}|^2 - |G_{ki}|^2 + 2|G_{ij}||G_{jk}||G_{ki}|\cos\Phi_h
\label{eq:detGh}
\end{equation}
where $\Phi_h = \phi_{ij} + \phi_{jk} + \phi_{ki}$ is the \emph{holonomy} around the triangle --- the gauge-invariant lattice field strength (Sec.~\ref{sec:holonomy}).

The dihedral angle at hinge $h$ is computed from the two tetrahedra sharing $h$, and the deficit angle $\delta_h = 2\pi - \sum_{\sigma \ni h} \theta_\sigma$ measures curvature. The Regge action is:
\begin{equation}
\boxed{S = \sum_h A_h\,\delta_h}
\label{eq:Regge}
\end{equation}
which converges to the Einstein--Hilbert action $\int R\sqrt{g}\,d^4x$ in the continuum limit \cite{Regge}.

\begin{remark}[No coupling constant in the action]
The action (\ref{eq:Regge}) contains no adjustable parameters: no $G_\text{Newton}$, no $\hbar$, no $\alpha$. It is pure geometry. All coupling constants emerge from the \emph{structure} of $G$, not from coefficients in the action (see Chapter~\ref{ch:couplings}).
\end{remark}

\subsection{Gauge connection from phase}
\label{sec:gauge_connection}

The phase $\phi_{ij} = \arg\langle\psi_i|\psi_j\rangle$ transforms under local rephasing $\psi_i \to e^{i\alpha_i}\psi_i$ as:
\begin{equation}
\phi_{ij} \to \phi_{ij} + \alpha_j - \alpha_i
\end{equation}
which is precisely the lattice gauge transformation $A_{ij} \to A_{ij} + \partial_j\alpha - \partial_i\alpha$. Thus $\phi_{ij}$ is a \emph{lattice gauge connection}.

\subsection{Holonomy as field strength}
\label{sec:holonomy}

The holonomy around a triangle $\{i, j, k\}$:
\begin{equation}
\Phi_{ijk} = \phi_{ij} + \phi_{jk} + \phi_{ki} = \arg(G_{ij}\,G_{jk}\,G_{ki})
\end{equation}
is gauge-invariant ($\alpha_i$ cancels cyclically). This is the discrete analogue of the field strength $F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu + [A_\mu, A_\nu]$.

From Eq.~(\ref{eq:detGh}), the hinge area $A_h = \sqrt{\det(G_h)}$ depends on $\Phi_h$: geometry and gauge are coupled through the determinant. This coupling is Einstein's equation in discrete form: matter (gauge fields) curves spacetime (hinge areas).

\subsection{Entanglement, Bell violation, and ER\,=\,EPR}

The polar decomposition $G_{ij} = |G_{ij}|\,e^{i\phi_{ij}}$ separates each element of $G$ into two independent components:
\begin{equation}
\underbrace{|G_{ij}|^2/d = W_{ij}}_{\text{geometry (distance)}} \qquad \underbrace{\phi_{ij} = \arg(G_{ij})}_{\text{gauge (correlation)}}
\end{equation}

``Locality'' is defined by $W$: two vertices are ``far apart'' when $W_{ij} \approx 0$. But $G_{ij} \neq 0$ for generic $\psi$ in $\mathbb{C}^5$, regardless of $W_{ij}$. The phase $\phi_{ij}$ persists even when the magnitude vanishes. This is entanglement:
\begin{equation}
\text{Entanglement} = \phi_{ij} \neq 0 \quad\text{while}\quad W_{ij} \approx 0.
\end{equation}

\textbf{Bell violation.} Bell's theorem proves that no \emph{local} hidden-variable model can reproduce quantum correlations. In DRLT, ``local'' means ``depends only on $W$,'' which discards $\phi$. A local model sets $\phi_{ij} = 0$, losing the phase correlations that $G$ carries. Including $\phi$ (i.e., using the full $G$, not just $W$) reproduces the quantum correlations. Bell's inequality is violated because $G \in \mathbb{C}$, not $\mathbb{R}$.

\textbf{No-signaling.} $W_{ij}$ carries energy and information (it is the magnitude of the interaction). $\phi_{ij}$ carries correlation (it is the relative phase). These are independent components of the same complex number. When $W_{ij} \approx 0$: no signal can propagate (no energy transfer), but correlation persists (phase is nonzero). No-signaling and entanglement coexist because magnitude and phase are independent.

\textbf{ER\,=\,EPR.} In 2013, Maldacena and Susskind conjectured that wormholes (ER bridges, geometric connections) and entanglement (EPR correlations, quantum connections) are the same phenomenon. In DRLT this is immediate:
\begin{equation}
G_{ij} = \underbrace{|G_{ij}|}_{\text{ER (geometry)}} \times \underbrace{e^{i\phi_{ij}}}_{\text{EPR (entanglement)}}
\end{equation}
A wormhole is a region where $W_{ij}$ is large (strong geometric connection) despite spatial separation. Entanglement is a region where $\phi_{ij}$ is correlated despite $W_{ij} \approx 0$. Both are components of the same complex inner product $G_{ij}$. The ER\,=\,EPR conjecture is the statement that $|G_{ij}|$ and $e^{i\phi_{ij}}$ belong to the same number. This is not a conjecture; it is the definition of $\mathbb{C}$.

\textbf{Ghost channels.} The Gram matrix $G$ has 5 significant eigenvalues and $N - 5$ eigenvalues at $0^+$ (the ghost dimensions). These $0^+$ eigenvalues connect distant simplices through ``almost zero'' hinge determinants: $\det(G_h) = 0^+$ means the hinge exists but carries negligible energy. Phase correlations persist through these ghost channels even when $W \approx 0$. The ghost dimensions are the geometric substrate of entanglement. (These 0$^+$ eigenvalues are distinct from the $\det(G_h)$ geometric contributions to coupling constants discussed in Chapter~\ref{ch:ghosts}.)

\subsection{Physical identification: A = spatial, B = temporal}
\label{sec:causal_split}

The vertex labels $A$ and $B$ were introduced in Chapter~\ref{ch:simplex_geometry} as neutral combinatorial types. We now identify them physically.

The 5 vertices of the simplex split into spatial and temporal:
\begin{equation}
\underbrace{A_1,\,A_2,\,A_3}_{n_A = 3\;\text{(spatial)}} \quad \underbrace{B_1,\,B_2}_{n_B = 2\;\text{(temporal)}}
\end{equation}

This identification follows from the Lorentz signature (Sec.~\ref{sec:Lorentz}): chirality of the weak force requires $n_B = 2$. Among all $\text{SU}(n)$, only $\text{SU}(2)$ has a pseudo-real fundamental representation ($\mathbf{2} \cong \bar{\mathbf{2}}$ via the antisymmetric tensor $\epsilon_{ab}$), which is the algebraic condition for chiral fermions. This forces the temporal sector to be $\mathbb{C}^2$, and therefore $n_A = d - n_B = 3$.

With this identification, the geometric structures from Chapter~\ref{ch:simplex_geometry} acquire physical meaning: AAA hinges become the strong sector, AAB the electromagnetic/weak sector, ABB the weak sector, and the BBB impossibility becomes the absence of a triple-temporal closure condition.

\subsection{Speed of light}

The lattice speed of light is the ratio of temporal to spatial vertex densities per dimension:
\begin{equation}
\boxed{c = \frac{n_B / d_B}{n_A / d_A} = \frac{2/1}{3/3} = 2}
\label{eq:speed_of_light}
\end{equation}
where $d_B = 1$ (time has 1 macroscopic dimension) and $d_A = 3$ (space has 3).

Every factor of $\sqrt{2}$ in the Standard Model --- $m = yv/\sqrt{2}$, the Higgs normalization, etc.\ --- is $\sqrt{c}$. The speed of light is not a fundamental constant but a \emph{lattice ratio}: time has twice the vertex density of space, per dimension.

\begin{remark}
The constancy of $c$ follows from its being a ratio of integers. It cannot vary, because $n_A = 3$ and $n_B = 2$ are fixed by the uniqueness of $d = 5$ (Theorem~\ref{thm:d5}).
\end{remark}

\subsubsection{From $c = 2$ to 299{,}792{,}458 m/s}

The lattice $c = 2$ is dimensionless: ``2 spatial cells per temporal tick.'' The SI value $c_\text{SI} = 299{,}792{,}458\;\text{m/s}$ is a unit conversion:
\begin{equation}
c_\text{SI} = c_\text{lattice} \times \frac{\Delta x}{\Delta t} = 2 \times \frac{\ell_P}{2\,t_P} = \frac{\ell_P}{t_P}
\end{equation}
where:
\begin{itemize}
\item $\Delta x = \ell_P = 1.616 \times 10^{-35}\;\text{m}$ (spatial lattice spacing $=$ Planck length).
\item $\Delta t = 2\,t_P = 1.078 \times 10^{-43}\;\text{s}$ (temporal lattice spacing $= 2 \times$ Planck time).
\end{itemize}

The temporal tick is $2\,t_P$ (not $t_P$) because $c = 2$: the temporal lattice is twice as coarse as Planck time. This is the lattice manifestation of $n_B = 2$ --- two temporal vertices per temporal dimension produce a spacing that is $c = 2$ times the ``natural'' (Planck) tick.

The numerator's 2 and denominator's 2 cancel: $c_\text{SI} = \ell_P/t_P$. This is the definition of Planck units --- but in DRLT, $\ell_P$ and $t_P$ are themselves derived from $\hbar = \sqrt{\det}/(4\ln 2)$ and the Regge action. The number 299{,}792{,}458 encodes the ratio of Planck length to Planck time, which encodes $\hbar$, $G$, and $c = 2$ --- all derived from $d = 5$.

\subsubsection{The status of $c$, $\hbar$, $G$}

The Regge action per hinge is $S_h/\hbar_h = (4G/c^3)\,\delta_h$. In the DRLT lattice:

\begin{center}
\begin{tabular}{lcll}
\hline
Constant & DRLT value & Status & Role \\
\hline
$c$ & 2 & \textbf{Derived} & Physics: $(n_B/d_B)/(n_A/d_A)$ \\
$\hbar$ & 1 & \textbf{Definition} & Fixes lattice units: $\hbar_h = A_h/(4\ln 2)$ \\
$G$ & $6.674 \times 10^{-11}$ & \textbf{Unit conversion} & Translates lattice $\leftrightarrow$ SI \\
\hline
\end{tabular}
\end{center}

$c = 2$ has physical content (the spacetime structure of $\mathbb{C}^5$). $\hbar = 1$ is a choice of units (it defines what ``1 lattice area'' means). $G$ encodes how large SI units (meters, kilograms, seconds) are compared to the lattice --- it is the statement that humans are $\sim\!10^{20}$ times larger than atoms, which are $\sim\!10^{15}$ times larger than the Planck scale. The ``weakness'' of gravity ($G \sim 10^{-11}$ in SI) is not a physical fact but a unit artifact: $G = 1$ in lattice units.

\textbf{Independent fundamental constants: zero.} $c$ is derived, $\hbar$ is definitional, $G$ is a unit conversion. The universe has no free parameters.

\subsection{Gauge group from chirality}

The full symmetry of $\mathbb{C}^5$ is $\text{SU}(5)$. The causal split breaks this:
\begin{equation}
\text{SU}(5) \;\to\; \text{SU}(3) \times \text{SU}(2) \times \text{U}(1)
\end{equation}

This is the Georgi--Glashow decomposition \cite{GeorgiGlashow}, but here it follows from the axiom rather than being postulated. The three factors are:
\begin{itemize}
\item $\text{SU}(3)$: internal rotations of $\mathbb{C}^3$ (strong force, color),
\item $\text{SU}(2)$: internal rotations of $\mathbb{C}^2$ (weak force, isospin),
\item $\text{U}(1)$: relative phase between $\mathbb{C}^3$ and $\mathbb{C}^2$ (electromagnetic force).
\end{itemize}

\subsection{Weinberg angle}

In $\text{SU}(5)$, the weak mixing angle at the unification scale is fixed by the trace condition on the generators:
\begin{equation}
\sin^2\theta_W = \frac{\text{Tr}(T_3^2)}{\text{Tr}(Q^2)} = \frac{3}{8}
\end{equation}
where $T_3$ and $Q$ are evaluated in the fundamental $\mathbf{5}$ representation. This is the standard Georgi--Glashow result, but here it is a consequence of $\mathbb{C}^5 = \mathbb{C}^3 \oplus \mathbb{C}^2$.

\subsection{Hinge classification and force hierarchy}

\subsubsection{Face-level structure}

Chapter~\ref{ch:simplex_geometry} showed that each face contains 4 hinges. The face type determines which hinge types participate:

\begin{center}
\begin{tabular}{lcccc}
\hline
Face type & AAA & AAB & ABB & Total \\
\hline
AAAB face ($\times 2$) & 1 & 3 & 0 & 4 \\
AABB face ($\times 3$) & 0 & 2 & 2 & 4 \\
\hline
\end{tabular}
\end{center}

Through an AAAB face, the AAA hinge participates: the full A-sector (strong force) is transmitted. Through an AABB face, no AAA hinge is involved: only the mixed sectors (electromagnetic, weak) are active.

\subsubsection{Hinge types}

The 10 hinges of the 4-simplex classify by vertex content (Chapter~\ref{ch:simplex_geometry}, now with physical identification):

\begin{center}
\begin{tabular}{lccl}
\hline
Type & Count & $\det(G_h)$ & Force \\
\hline
AAA (3 A-vertices) & $\binom{3}{3}\binom{2}{0} = 1$ & $> 0$ & Strong \\
AAB (2 A + 1 B) & $\binom{3}{2}\binom{2}{1} = 6$ & $> 0$ & Electromagnetic / Weak \\
ABB (1 A + 2 B) & $\binom{3}{1}\binom{2}{2} = 3$ & $= 0$ always & Classical ($n_B = 2 < 3$) \\
BBB (3 B-vertices) & $\binom{2}{3} = 0$ & --- & Impossible \\
\hline
\end{tabular}
\end{center}

The ABB hinges have $\det(G_h) = 0$ because $n_B = 2 < 3$: two temporal vertices cannot span a nondegenerate triangle. BBB is impossible since $\binom{2}{3} = 0$.

This gives a natural force hierarchy: 7 quantum hinges (AAA + AAB) mediate the strong and electroweak forces, while ABB hinges are degenerate ($\hbar \to 0$). The weak force has no dedicated hinge (BBB $= 0$) and must ``borrow'' from the AAB sector, providing a geometric explanation for its weakness.

\subsection{Why exactly four forces}

The number of gauge forces equals the number of ways to distribute 3 hinge vertices into the A-sector ($\mathbb{C}^3$) and B-sector ($\mathbb{C}^2$), excluding the impossible case:
\begin{equation}
\text{Gauge forces} = |\{(3,0),\;(2,1),\;(1,2)\}| = 3. \qquad (0,3)\;\text{is forbidden since } n_B = 2 < 3.
\end{equation}
Gravity adds one more: it operates not \emph{within} a hinge but \emph{between} hinges (the deficit angle $\delta_h$ measures how neighboring hinges are arranged). Total:
\begin{equation}
\text{Forces} = \underbrace{\min(3, n_B) + 1}_{\text{gauge + gravity}} = 3 + 1 = 4.
\end{equation}

A 5th force is mathematically forbidden: it would require either a 4th hinge type (impossible since $n_B = 2$) or a 2nd geometric level beyond hinges (but hinge-to-hinge is already the simplicial boundary structure, and there is nothing beyond it).

\begin{center}
\begin{tabular}{lccl}
\hline
Force & Level & Hinge type & Metaphor \\
\hline
Strong & within hinge & AAA & The room \\
EM & within hinge & AAB & The glue \\
Weak & within hinge & ABB & The door \\
Gravity & between hinges & $\det(G_h)$ & The road \\
\hline
\end{tabular}
\end{center}

\subsection{Fermions and bosons from vertex selection}
\label{sec:spin_statistics}

Chapter~\ref{ch:simplex_geometry} derived 15 antisymmetric vertex selection patterns purely from combinatorics. We now identify them with the Standard Model fermions.

\subsubsection{The $\bar{5}$: 1-vertex selections}

Selecting 1 vertex from 5 gives $\binom{5}{1} = 5$ patterns. With the physical identification $A$ = spatial, $B$ = temporal:

\begin{center}
\begin{tabular}{cccl}
\hline
Selection & Fingerprint & Count & Standard Model \\
\hline
A vertex & AAA$\times$1 + AAB$\times$4 + ABB$\times$1 & 3 & $d_R$ (3 colors) \\
B vertex & AAB$\times$3 + ABB$\times$3 & 2 & $(\nu, e)_L$ doublet \\
\hline
\multicolumn{3}{r}{Total} & 5 \\
\hline
\end{tabular}
\end{center}

A-type selections participate in the AAA hinge $\to$ confined. B-type selections do not $\to$ free.

\subsubsection{The $10$: 2-vertex selections}

Selecting 2 vertices gives $\binom{5}{2} = 10$ antisymmetric patterns:

\begin{center}
\begin{tabular}{cccl}
\hline
Pair type & Fingerprint & Count & Standard Model \\
\hline
AA & AAA$\times$1 + AAB$\times$2 & $\binom{3}{2} = 3$ & $u_R$ (3 colors) \\
AB & AAB$\times$2 + ABB$\times$1 & $3 \times 2 = 6$ & $(u, d)_L$ doublet ($3 \times 2$) \\
BB & ABB$\times$3 & $\binom{2}{2} = 1$ & $e_R$ singlet \\
\hline
\multicolumn{3}{r}{Total} & 10 \\
\hline
\end{tabular}
\end{center}

\subsubsection{Total: 15 Weyl fermions per generation}

$\binom{5}{1} + \binom{5}{2} = 5 + 10 = 15$. This is the exact content of one Standard Model generation under $\text{SU}(5)$: $\bar{\mathbf{5}} \oplus \mathbf{10}$.

\textbf{Spin-$1/2$:} The B-sector is $\mathbb{C}^2$. $\text{SU}(2)$ acts on $\mathbb{C}^2$ as the spin-$1/2$ representation. The ``spin'' of a fermion is the dimension of the temporal sector it occupies.

\textbf{Pauli exclusion:} Two identical vertices $\psi_i = \psi_j$ merge into one point ($\det = 0$). The simplex loses a vertex, destroying information. This is Pauli exclusion from geometry.

\subsubsection{Bosons: symmetric structures}

The symmetric structures are edges (inner products $G_{ij} = \langle\psi_i|\psi_j\rangle$), which carry 2 indices. In $\text{SU}(2)$: $\frac{1}{2} \otimes \frac{1}{2} = 0 \oplus 1$, giving spin-0 and spin-1.

\textbf{Gravity} is not a vertex or edge. It is the deficit angle $\delta_h = 2\pi - \sum\theta$ --- the arrangement of simplices around a hinge, not an object on a simplex. The ``spin 2'' label reflects the 2-dimensional nature of the hinge.

\begin{remark}[Spin-statistics in one line]
1-vertex and 2-vertex selections are antisymmetric $\to$ Fermi. Edges ($G_{ij} = G_{ji}^*$) are Hermitian-symmetric $\to$ Bose.
\end{remark}

\subsection{Singularity resolution}

\begin{proposition}
No two distinct points can have $ds^2 = 0$.
\end{proposition}

\begin{proof}
$ds^2 = 0$ requires $|G_{ij}| = 1$, i.e., $\psi_i = e^{i\theta}\psi_j$, which means $i$ and $j$ represent the same physical point in $\mathbb{CP}^4$. Distinct points always have $ds^2 > 0$.
\end{proof}

This eliminates singularities: gravitational collapse drives $\psi_i \to \psi_j$ (alignment), giving $\det(G_h) \to 0$, hence $\hbar_\text{eff} \to 0$ (Chapter~\ref{ch:hbar}). The region becomes classical vacuum, not a singularity. The ``center'' of a black hole is where $\psi$ values align --- it is vacuum, not infinite density.

More precisely, $\det(G_h) = 0$ is a codimension-2 condition in the configuration space of $\psi$-vectors (requiring 3 vectors to span a 2-dimensional subspace).  It has measure zero and is dynamically unstable: any perturbation restores $\det > 0$.  Singularities cannot be \emph{maintained}, though they can be transiently approached.

%---------------------------------------------------------------------
\subsection{Shadow theorem: spacetime as moment map image}
\label{sec:shadow}
%---------------------------------------------------------------------

Since $\psi_i$ and $e^{i\theta}\psi_i$ define the same physical state, the true state space per vertex is $\mathbb{CP}^4$, a toric variety of real dimension~8.

\begin{proposition}[Toric decomposition of $\mathbb{CP}^4$]
\label{prop:toric}
The standard torus $T^4$ acts on $\mathbb{CP}^4$ by
\begin{equation}
(\theta_1,\ldots,\theta_4)\cdot[z_0:\cdots:z_4]
= [z_0 : e^{i\theta_1}z_1 : \cdots : e^{i\theta_4}z_4].
\end{equation}
The moment map $\mu: \mathbb{CP}^4 \to \Delta^4$ is $\mu([z]) = (|z_0|^2/|z|^2,\ldots,|z_4|^2/|z|^2)$, whose image is the probability simplex $\Delta^4$.  The generic fiber is $\mu^{-1}(p) \cong T^4$.
\end{proposition}

This is a standard result in toric geometry~\cite{fulton1993}.  We interpret it as follows.

\begin{itemize}
\item \textbf{Base $\Delta^4$} (4 real dim): carries the modulus information $|G_{ij}|$.  This determines distances ($ds^2 = 1 - d\,W_{ij}$, where $W = |G|^2/d$), deficit angles, and curvature.  A theory built from $|G|$ alone is general relativity.

\item \textbf{Fiber $T^4$} (4 real dim): carries the phase information $\arg(G_{ij})$.  This determines holonomies, gauge connections, and field strengths.  A theory built from $\arg(G)$ alone is gauge theory (the Standard Model).

\item \textbf{Total $\mathbb{CP}^4$} (8 real dim): carries both modulus and phase simultaneously through $G_{ij} \in \mathbb{C}$.  This is the DRLT framework.
\end{itemize}

\noindent
\textit{Spacetime as observed is the moment map image (the ``shadow'') $\Delta^4$.  The gauge forces are the fiber $T^4$ that the projection forgets.}

%---------------------------------------------------------------------
\subsection{Gauge group from fiber decomposition}
\label{sec:toric_gauge}
%---------------------------------------------------------------------

\begin{theorem}[Maximal torus decomposition]
\label{thm:torus_decomposition}
The $(3,2)$ partition $\mathbb{C}^5 = \mathbb{C}^3 \oplus \mathbb{C}^2$ decomposes the fiber:
\begin{equation}
T^4 = T^2_{\mathrm{SU}(3)} \times T^1_{\mathrm{SU}(2)} \times T^1_{\mathrm{U}(1)}.
\end{equation}
\end{theorem}

\begin{proof}
$\mathrm{SU}(3)$ acts on $\mathbb{C}^3$; its maximal torus has rank~2 (two independent relative phases among $z_0, z_1, z_2$).  $\mathrm{SU}(2)$ acts on $\mathbb{C}^2$; its maximal torus has rank~1 (one relative phase, $z_3$ vs $z_4$).  $\mathrm{U}(1)_Y$ is the relative phase between the $\mathbb{C}^3$ and $\mathbb{C}^2$ sectors.  Dimensions: $2 + 1 + 1 = 4 = \dim T^4$.
\end{proof}

Since the gauge group is reconstructed from its maximal torus via the root system, this proves that $\mathrm{SU}(3) \times \mathrm{SU}(2) \times \mathrm{U}(1)$ is not postulated but \emph{derived} from the toric structure of $\mathbb{CP}^4$ and the $(3,2)$ split.

\begin{corollary}[Symmetry breaking is topologically inevitable]
\label{cor:breaking}
$\Delta^4$ is compact, so $\partial\Delta^4 \neq \varnothing$.  On $\partial\Delta^4$, some $|z_k|^2 = 0$ and the corresponding fiber direction degenerates.  Therefore the full gauge symmetry \emph{must} be broken somewhere --- symmetry breaking is a topological necessity, not a dynamical accident.
\end{corollary}

%---------------------------------------------------------------------
\subsection{Metric--gauge inseparability}
\label{sec:inseparability}
%---------------------------------------------------------------------

The hinge determinant (Eq.~\ref{eq:det_Gh} in Chapter~\ref{ch:simplex_geometry}) contains both modulus and phase:
\begin{equation}
\det(G_h) = 1 - |G_{ij}|^2 - |G_{jk}|^2 - |G_{ki}|^2 + 2\,|G_{ij}|\,|G_{jk}|\,|G_{ki}|\cos\Phi_h,
\end{equation}
where $\Phi_h = \arg(G_{ij} G_{jk} G_{ki})$ is the holonomy around the triangle.  The last term couples metric ($|G|$) and gauge ($\Phi$) information \emph{multiplicatively} inside the determinant.

Consequence: the Regge action $S = \sum_h A_h\,\delta_h = \sum_h \sqrt{\det(G_h)}\;\delta_h(G)$ depends on both metric and gauge degrees of freedom through the same determinant.  They cannot be separated.  This inseparability is the mathematical content of gravity--gauge unification in the simplex framework.
