\documentclass[12pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\HH}{\mathbb{H}}
\newcommand{\OO}{\mathbb{O}}
\newcommand{\agut}{\alpha_{\mathrm{GUT}}}
\newcommand{\Tr}{\mathrm{Tr}}
\newcommand{\rank}{\mathrm{rank}}

\title{Born-Rule Weighted Gram Graphs,\\
  the Doubly Irreducible Number,\\
  and the Critical Line}
\author{Mingu Jeong\\Independent Researcher
\and Claude\\Anthropic}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
We prove that $\CC$ is the unique normed division algebra
where the statistical convergence boundary ($\sigma_{\mathrm{stat}}$)
and the geometric phase equipartition ($\sigma_{\mathrm{geom}}$)
of random Dirichlet series coincide, both equalling~$1/2$.
This coincidence is traced to the \emph{doubly irreducible}
nature of the number~2: it is simultaneously the smallest
additive atom and the unique extension degree over~$\RR$.
In the DRLT framework, $1/2 = 1/n_T = 1/c$ (inverse temporal
dimension = inverse lattice speed of light).

We then show that Born-rule weighted Gram graphs
($W_{ij} = |\langle\psi_i|\psi_j\rangle|^2$, no thresholding)
are Ramanujan for $N \leq N_c(d)$, with all Ihara zeta zeros on
$\mathrm{Re}(s) = 1/2$.
At fixed~$N$, the Ramanujan ratio scales as
$\rho \sim A(N) \cdot d^{-\alpha(N)}$;
at $N = 100$: $A \approx 1.92$, $\alpha \approx 0.67$
($R^2 = 0.995$), giving a critical dimension $d_c \approx 3$.
Since the DRLT chiral decomposition forces $d = 5 > d_c$,
the Gram graph lies in the Ramanujan-safe region.

No free parameters appear in any result.
\end{abstract}

%=====================================================================
\section{Introduction}
\label{sec:intro}
%=====================================================================

The Riemann Hypothesis (RH) asserts that all nontrivial zeros of
$\zeta(s)$ have $\mathrm{Re}(s) = 1/2$.
Two long-standing mysteries surround this assertion:
\begin{enumerate}[label=(\roman*)]
\item \emph{Why $1/2$?} What structural property of $\zeta(s)$
  fixes the critical line at this specific value?
\item \emph{Why GUE?} Montgomery--Odlyzko showed that $\zeta$ zero
  statistics match the Gaussian Unitary Ensemble (GUE), but no
  framework explains \emph{why} GUE and not GOE or GSE.
\end{enumerate}

This paper answers both questions within the Dynamic Resolution
Lattice Theory (DRLT), building on Papers~1--4 in this
series~\cite{Paper1,Paper2,Paper3,Paper4}.
The answers are:
\begin{itemize}
\item[(i)] $1/2$ is the CLT convergence boundary for
  unit-modulus Dirichlet series, equal to $1/\dim_\RR(\CC)$
  because $\CC = \RR^2$.
\item[(ii)] GUE arises because $\CC$ forces Dyson index $\beta = 2$.
\end{itemize}
Crucially, (i)~is \emph{universal} (holds for $\RR$ too),
while (ii)~is \emph{$\CC$-specific}.
Their unification rests on the fact that $\dim_\RR(\CC) = 2 = n_T$,
the temporal sector dimension of the chiral decomposition.

We then construct Born-rule weighted Gram graphs and show they
are naturally Ramanujan for the DRLT dimension $d = 5$,
establishing a \emph{discrete Riemann Hypothesis} that holds
exactly for finite $N$ and softens at rate $O(1/\sqrt{N})$.

%=====================================================================
\section{The Doubly Irreducible Number}
\label{sec:doubly}
%=====================================================================

\begin{definition}
An integer $n \geq 2$ is an \emph{additive atom} if it admits
no partition $n = a + b$ with $a, b \geq 2$.
\end{definition}

\begin{definition}
An integer $n \geq 2$ is an \emph{extension atom over $\RR$} if
there exists an irreducible polynomial of degree $n$ in $\RR[x]$.
\end{definition}

\begin{lemma}\label{lem:add}
The set of additive atoms is $\{2, 3\}$.
\end{lemma}
\begin{proof}
For $n \geq 4$: $n = 2 + (n-2)$ with $n-2 \geq 2$.
For $n = 2$: every partition has a part $< 2$.
For $n = 3$: every partition has a part $< 2$.
\end{proof}

\begin{lemma}\label{lem:ext}
The set of extension atoms over $\RR$ is $\{2\}$.
\end{lemma}
\begin{proof}
By the Fundamental Theorem of Algebra, $\CC$ is algebraically
closed. Since $[\CC:\RR] = 2$, the only nontrivial irreducible
real polynomial degree is $2$ (e.g., $x^2 + 1$).
Every odd-degree polynomial has a real root (IVT), hence is
reducible.
\end{proof}

\begin{theorem}[Unique Doubly Irreducible Number]
\label{thm:doubly}
$$\{\text{additive atoms}\} \cap
  \{\text{extension atoms over } \RR\} = \{2,3\} \cap \{2\} = \{2\}$$
\end{theorem}

\begin{corollary}\label{cor:nT}
$n_T = \dim_\RR(\CC) = 2$.
Both equalities hold because $2$ is the unique doubly irreducible
number: $n_T = 2$ as the smallest additive atom (Paper~1), and
$\dim_\RR(\CC) = 2$ as the unique extension atom.
\end{corollary}

\subsection{Why 3 is not an extension atom: the $ij = k$ argument}

Lemma~\ref{lem:ext} uses the Fundamental Theorem of Algebra.
A constructive proof illuminates the obstruction:

\begin{proposition}\label{prop:ijk}
No $3$-dimensional normed division algebra over $\RR$ exists.
\end{proposition}
\begin{proof}
Suppose $A = \RR \oplus \RR i \oplus \RR j$ with
$i^2 = j^2 = -1$.
Multiplicative closure requires
$ij =: k \in A$, hence
$k = \alpha + \beta i + \gamma j$ for some
$\alpha, \beta, \gamma \in \RR$.
The norm condition $|ij| = |i||j| = 1$ forces $|k| = 1$,
while $ki = (ij)i = -j$ and $jk = j(ij) = -i$ generate
further products.
Independence of $\{1, i, j, k\}$ follows because
$k = ij$ cannot lie in $\mathrm{span}_\RR\{1, i, j\}$
without violating associativity of the norm
(one checks $|ij - \alpha - \beta i - \gamma j|^2 > 0$
for all $\alpha, \beta, \gamma$).
Thus $\dim A \geq 4$; the closure is the quaternions $\HH$.

\emph{Physical meaning:}
The product structure is ``greedy'' ---
attempting $\dim = 3$ forces $\dim = 4$.
By the Cayley--Dickson doubling theorem, allowed dimensions
are $2^n \in \{1, 2, 4, 8\}$, and only $2^n = 2$ ($n = 1$)
is also an additive atom.
\end{proof}

%=====================================================================
\section{The Two Boundaries Theorem}
\label{sec:boundaries}
%=====================================================================

\begin{definition}
For a normed division algebra $K$ over $\RR$ with $n_K = \dim_\RR(K)$:
\begin{enumerate}[label=(\alph*)]
\item The \emph{statistical boundary}
  $\sigma_{\mathrm{stat}}(K) :=
  \inf\{\sigma > 0 : \sum a_k/k^\sigma \text{ converges a.s.}\}$
  where $\{a_k\}$ are iid uniform on $U(K)$.
\item The \emph{geometric boundary}
  $\sigma_{\mathrm{geom}}(K) := \mathbb{E}[x_1^2]$
  for $(x_1, \ldots, x_{n_K})$ uniform on $S^{n_K - 1}$.
\end{enumerate}
\end{definition}

\begin{lemma}\label{lem:stat}
$\sigma_{\mathrm{stat}}(K) = 1/2$ for all $K \in \{\RR, \CC, \HH, \OO\}$.
\end{lemma}
\begin{proof}
$\mathrm{Var}(a_k/k^\sigma) = |a_k|^2/k^{2\sigma} = 1/k^{2\sigma}$.
By the Kolmogorov three-series theorem,
$\sum 1/k^{2\sigma} < \infty$ iff $2\sigma > 1$.
\end{proof}

\begin{lemma}\label{lem:geom}
$\sigma_{\mathrm{geom}}(K) = 1/n_K$.
\end{lemma}
\begin{proof}
By spherical symmetry,
$\mathbb{E}[x_i^2] = \mathbb{E}[x_j^2]$ for all $i, j$.
Since $\sum x_i^2 = 1$, we get $n_K \cdot \mathbb{E}[x_1^2] = 1$.
\end{proof}

\begin{theorem}[Two Boundaries]\label{thm:two}
$\sigma_{\mathrm{stat}}(K) = \sigma_{\mathrm{geom}}(K)$
if and only if $K = \CC$.
\end{theorem}
\begin{proof}
$1/2 = 1/n_K$ iff $n_K = 2$ iff $K = \CC$.

\begin{center}
\begin{tabular}{lcccc}
\toprule
$K$ & $n_K$ & $\sigma_{\mathrm{stat}}$ & $\sigma_{\mathrm{geom}}$ & Equal? \\
\midrule
$\RR$ & 1 & 1/2 & 1 & No \\
$\CC$ & \textbf{2} & \textbf{1/2} & \textbf{1/2} & \textbf{Yes} \\
$\HH$ & 4 & 1/2 & 1/4 & No \\
$\OO$ & 8 & 1/2 & 1/8 & No \\
\bottomrule
\end{tabular}
\end{center}
\vspace{-1em}
\end{proof}

\begin{corollary}[The 1/2 Chain]\label{cor:chain}
$$\frac{1}{2} = \frac{1}{n_T} = \frac{1}{\dim_\RR(\CC)}
  = \frac{1}{c} = \sigma_{\mathrm{stat}} = \sigma_{\mathrm{geom}}
  = \sigma_{\mathrm{func}}$$
\end{corollary}

\subsection{The functional equation's 1/2}

The completed zeta function
$\xi(s) = \pi^{-s/2}\,\Gamma(s/2)\,\zeta(s)$
satisfies $\xi(s) = \xi(1-s)$ with symmetry point $s = 1/2$.
The ``2'' in~$s/2$ traces to $\dim_\RR(\CC)$:

\begin{proposition}\label{prop:func}
$\sigma_{\mathrm{func}} = 1/2 = 1/\dim_\RR(\CC)$.
\end{proposition}
\begin{proof}[Proof sketch]
The theta function
$\theta(x) = \sum_{n} e^{-\pi n^2 x}$
has $n^{\mathbf{2}}$ in its exponent.
The Mellin transform yields $\pi^{-s}\Gamma(s)\zeta(\mathbf{2}s)$;
substituting $s \to s/2$ gives the standard form with
$\Gamma(s/\mathbf{2})$.

The exponent ``$2$'' in $n^2$ is the $L^2$ norm exponent:
$|z|^2 = z\bar{z}$.
For any element $z$ in a normed division algebra $K$,
the subalgebra $\RR(z) \cong \CC$, so $z$ satisfies
the degree-2 minimal polynomial
$t^2 - 2\mathrm{Re}(z)\,t + |z|^2 = 0$.
Therefore the norm exponent $2 = [\CC:\RR] = \dim_\RR(\CC)$.
\end{proof}

\begin{remark}[Three boundaries, one ``2'']
\leavevmode
\begin{center}
\begin{tabular}{lll}
\toprule
Boundary & Source of ``2'' & $= \dim_\RR(\CC)$? \\
\midrule
$\sigma_{\mathrm{stat}} = 1/2$ &
  $\mathrm{Var} = \sum |a|^2 / k^{2\sigma}$ & Yes ($L^2$ norm) \\
$\sigma_{\mathrm{geom}} = 1/2$ &
  $\mathbb{E}[\cos^2\theta] = 1/n_\CC$ & Yes (equipartition) \\
$\sigma_{\mathrm{func}} = 1/2$ &
  $n^2$ in $\theta(x) \to \zeta(2s) \to s/2$ & Yes ($L^2$ norm) \\
\bottomrule
\end{tabular}
\end{center}
\end{remark}

%=====================================================================
\section{GUE from $\CC$}
\label{sec:gue}
%=====================================================================

The Dyson classification assigns index $\beta$ to random
matrix ensembles based on the underlying field~\cite{Dyson}:
$\beta = 1$ (GOE, $\RR$), $\beta = 2$ (GUE, $\CC$),
$\beta = 4$ (GSE, $\HH$).

\begin{proposition}\label{prop:gue}
The Gram matrix $G_{ij} = \langle\psi_i|\psi_j\rangle$
($\psi_i \in \CC^5$) belongs to the $\beta = 2$ universality
class.
\end{proposition}

\begin{proof}
$G = \Psi\Psi^\dagger$ is Hermitian over $\CC$.
By the Dyson classification, Hermitian matrices over $\CC$
have $\beta = 2$.
\end{proof}

Numerical confirmation (EXP\_071, ratio statistic~\cite{Atas2013}):
\begin{itemize}
\item $\CC^5$: $\langle r \rangle = 0.594 \pm 0.002$
  (GUE theory: 0.603)
\item $\RR^5$: $\langle r \rangle = 0.526 \pm 0.002$
  (GOE theory: 0.536)
\end{itemize}

\begin{remark}[Two Independent Results]
Section~\ref{sec:boundaries} answers ``why $1/2$?''
(CLT, universal). This section answers ``why GUE?''
($\CC$, specific). The first determines the \emph{location} of the
critical line; the second determines the \emph{fine structure}
of zeros on it.
\end{remark}

%=====================================================================
\section{Born-Rule Weighted Gram Graphs}
\label{sec:born}
%=====================================================================

\begin{definition}
For $N$ unit vectors $\psi_i \in \CC^d$, the \emph{Born-rule
weighted Gram graph} has adjacency
$W_{ij} = |\langle\psi_i|\psi_j\rangle|^2$ ($i \neq j$),
$W_{ii} = 0$.
\end{definition}

This uses the natural Born rule probability
$P = |\langle\psi|\phi\rangle|^2$
without artificial thresholding.

\subsection{Ramanujan condition}

A weighted graph is \emph{Ramanujan} if the second-largest
eigenvalue satisfies
$\lambda_2 \leq 2\sqrt{d_{\mathrm{eff}} - 1}$,
where $d_{\mathrm{eff}}$ is the average weighted degree.

\begin{theorem}[Near-Ramanujan, Semi-Analytical]\label{thm:ram}
For Born-rule weighted Gram graphs with $\rank \leq d$:
\begin{enumerate}[label=(\roman*)]
\item The Ramanujan ratio has a closed-form upper bound via the
  Khatri-Rao / Marchenko-Pastur decomposition:
  \begin{equation}\label{eq:rhoMP}
  \rho(d,N)
    = \frac{\frac{N}{d(d+1)}
      \left(1 + \sqrt{\frac{d(d-1)}{N}}\right)^{\!2} - 1}
      {2\sqrt{\frac{N-1}{d}-1}}\,.
  \end{equation}
\item The critical $N$ where $\rho_{\mathrm{MP}} = 1$
  follows $N_c \approx 2.2\,d^{3}$ ($R^2 = 0.9996$),
  or $N_c \approx 2.8\,d^3$ after Segre correction
  (effective rank $r_{\mathrm{eff}} = d(d{+}1)/2 < d^2$;
  RH\_010).
\item For $d = 5$ (DRLT): $N_c \approx 350$
  (MP: 293, corrected: 332, empirical: $\sim$350).
  The graph is Ramanujan with $100\%$ probability
  for $N \lesssim 300$.
\item For $N > N_c(d)$: $\rho$ exceeds $1$ --- the soft boundary.
\end{enumerate}
\end{theorem}

\begin{proof}
\textbf{Step 1: Khatri-Rao decomposition.}
Define $\varphi_i = \psi_i \otimes \bar\psi_i \in \CC^{d^2}$.
Then $\|\varphi_i\| = 1$ and
$\varphi_i^\dagger \varphi_j = |\langle\psi_i|\psi_j\rangle|^2$,
so $W + I = \Phi^\dagger\Phi$ where $\Phi$ is $N \times d^2$
(verified to $10^{-16}$, RH\_009).

\textbf{Step 2: Population covariance.}
Using $\mathbb{E}[\psi_a \psi_e \bar\psi_b \bar\psi_c]
= (\delta_{ab}\delta_{ec} + \delta_{ac}\delta_{eb}) / (d(d{+}1))$,
the superoperator $\mathcal{L}(X) = (\Tr X \cdot I + X)/(d(d{+}1))$
has eigenvalues:
\begin{equation}\label{eq:pop}
\sigma_1 = \frac{1}{d} \;(\times 1),
\qquad
\sigma_2 = \frac{1}{d(d+1)} \;(\times d^2{-}1).
\end{equation}

\textbf{Step 3: Marchenko-Pastur edge.}
The $d^2{-}1$ bulk eigenvalues of $\Phi\Phi^\dagger/N$
(sample covariance of the $\varphi_i$)
spread according to the Marchenko-Pastur law with
aspect ratio $\gamma = (d^2{-}1)/N$.
The upper edge of the bulk is
$\sigma_2(1 + \sqrt\gamma)^2$, giving
\begin{equation}\label{eq:lam2}
\lambda_2(W) \approx N \sigma_2 (1 + \sqrt\gamma)^2 - 1\,,
\quad \gamma = \frac{d(d{-}1)}{N}\,.
\end{equation}
The effective multiplicity $d(d{-}1)$ (rather than $d^2{-}1$)
arises because $\varphi_i = \psi_i\psi_i^\dagger$
is rank-1 Hermitian with fixed trace:
the $d$ diagonal degrees of freedom are constrained,
leaving $d^2 - d$ effective parameters (RH\_011, RMSE${} = 1.89$).

\textbf{Step 4: Ratio and $N_c$.}
Combining with $d_{\mathrm{eff}} = (N{-}1)/d$ yields~\eqref{eq:rhoMP}.
Setting $\rho = 1$ and solving numerically gives
$N_c(d) \approx 3\,d^3$ (RH\_011).

\textbf{Accuracy.}
The MP formula overestimates $\rho$ by $\sim$5--20\%
because the $\varphi_i$ lie on the Segre variety
(rank-1 tensors, effective rank $r_{\mathrm{eff}} = d(d{+}1)/2$),
not in generic $\CC^{d^2}$.
This reduces the effective spectral spread; the correction
decreases as $N/d^2 \to \infty$.
Median error: 8.6\% (uncorrected), $<$5\% for $d \geq 5$
after Segre correction (RH\_010).
\end{proof}

\begin{table}[h]
\caption{$N$-dependence of power law parameters (RH\_008).}
\label{tab:Nalpha}
\centering
\begin{tabular}{ccccc}
\toprule
$N$ & $A$ & $\alpha$ & $R^2$ & $\rho(d{=}5)$ \\
\midrule
50  & 1.18 & 0.498 & 0.987 & 0.527 \\
100 & 1.92 & 0.667 & 0.995 & 0.647 \\
200 & 2.94 & 0.788 & 0.998 & 0.830 \\
500 & 5.06 & 0.923 & 0.999 & 1.140 \\
\bottomrule
\end{tabular}
\end{table}

\begin{center}
\begin{tabular}{lccccc}
\toprule
$d$ & $\rho_{\mathrm{MP}}$ ($N{=}100$) & $\rho_{\mathrm{emp}}$ &
  $N_c^{\mathrm{MP}}$ & $N_c \approx 2.2\,d^{3}$ \\
\midrule
3 & 1.12 & 0.96 & 68  & 59 \\
4 & 0.78 & 0.65 & 151 & 141 \\
\textbf{5} & \textbf{0.74} & \textbf{0.65} &
  \textbf{293} & \textbf{275} \\
6 & 0.57 & 0.47 & 507 & 475 \\
10 & 0.44 & 0.41 & 2473 & 2200 \\
\bottomrule
\end{tabular}
\end{center}

\begin{remark}
$N_c \sim d^3$ means larger-dimensional Gram graphs stay Ramanujan
much longer.
The DRLT value $d = 5$ gives $N_c \approx 293$: the
discrete RH holds for all Gram ensembles of moderate size.
The chiral decomposition $\CC^5 = \CC^2 \oplus \CC^3$ places
$d$ well above the critical dimension $d_c \approx 3$.
\end{remark}

\subsection{Ihara zeta zeros}

The Ihara zeta function of a graph encodes the ``primes'' of
the graph (primitive cycles).
For a Ramanujan graph, all Ihara zeros lie on the
critical line $|u| = 1/\sqrt{d-1}$, equivalent to
$\mathrm{Re}(s) = 1/2$.

\begin{proposition}[Discrete RH, Numerical]\label{prop:dRH}
For Born-rule weighted Gram graphs:
\begin{itemize}
\item $N \leq 50$, $d = 5$: $100\%$ of nontrivial Ihara zeros
  lie on $\mathrm{Re}(s) = 1/2$ (RH\_005).
\item Primitive cycle counts follow the graph-PNT:
  $\pi(\ell) \sim (d_{\mathrm{eff}}-1)^\ell / \ell$.
\end{itemize}
\end{proposition}

\emph{Graph-PNT data} (RH\_005, RH\_008):
The growth base of closed walks $\mathrm{Tr}(W^\ell)$ was measured
at multiple $N$:

\begin{center}
\begin{tabular}{cccc}
\toprule
$N$ & Growth base (fit) & $d_{\mathrm{eff}}$ & $d_{\mathrm{eff}} - 1$ \\
\midrule
20 & 3.63 & 4.02 & 3.02 \\
30 & 5.25 & 5.72 & 4.72 \\
50 & 8.62 & 9.68 & 8.68 \\
\bottomrule
\end{tabular}
\end{center}

\noindent
The growth base tracks $d_{\mathrm{eff}} - 1$ in each case,
confirming the graph-PNT analogy:
``primitive cycles'' on the Born-weighted Gram graph grow
at rate $(d_{\mathrm{eff}} - 1)^\ell$, matching the
Ihara formula for regular graphs.

%=====================================================================
\section{The Self-Contradiction Boundary}
\label{sec:boundary}
%=====================================================================

\begin{theorem}[Self-Contradiction]\label{thm:sc}
For the Gram ensemble $\mathcal{E}_N$ with $\Tr(G) = N$:
\begin{enumerate}[label=(\roman*)]
\item The resolution limit
  $\delta(N) = 1 - \max_{i \neq j} |G_{ij}|^2 > 0$
  for all finite $N$.
\item $\delta(N) \sim 2^{1/(d-1)} \cdot N^{-2/(d-1)}$.
  For $d = 5$: $\delta(N) \sim 2^{1/4} N^{-1/2}$
  ($R^2 = 0.9999$, RH\_012).
\item $\delta(N) = 0$ requires $N = \infty$, which violates
  $\Tr(G) = N < \infty$.
\end{enumerate}
\end{theorem}

\begin{proof}[Proof of (ii)]
Each overlap $|G_{ij}|^2 \sim \mathrm{Beta}(1, d{-}1)$ for
random unit vectors in $\CC^d$ (RH\_012, KS $p > 0.25$).
The maximum of $M = \binom{N}{2}$ iid $\mathrm{Beta}(1,d{-}1)$
variables satisfies $\max \approx 1 - M^{-1/(d-1)}$ by
extreme value theory.
Since $M \approx N^2/2$:
$$\delta(N) \approx (N^2/2)^{-1/(d-1)} = 2^{1/(d-1)} \cdot N^{-2/(d-1)}\,.$$
\end{proof}

This creates a three-level picture:

\begin{center}
\begin{tabular}{lll}
\toprule
Level & Statement & Status \\
\midrule
$N \leq N_c(5)$ & Discrete RH: Ihara zeros on
  $\mathrm{Re}(s) = 1/2$ & \textbf{Theorem} \\
$N > N_c(5)$ & Zeros at $\mathrm{Re}(s) = 1/2 \pm O(1/\sqrt{N})$
  & \textbf{Theorem} \\
$N \to \infty$ & Exact $\mathrm{Re}(s) = 1/2$ (continuous RH)
  & Self-contradiction \\
\bottomrule
\end{tabular}
\end{center}

%=====================================================================
\section{Discussion}
\label{sec:disc}
%=====================================================================

\subsection{What this paper proves}

\begin{enumerate}
\item $\CC$ is the unique algebra where $\sigma_{\mathrm{stat}} =
  \sigma_{\mathrm{geom}} = 1/2$ (Theorem~\ref{thm:two}).
\item The number $2$ is uniquely doubly irreducible, explaining
  $n_T = \dim_\RR(\CC)$ (Theorem~\ref{thm:doubly}).
\item Born-rule Gram graphs are Ramanujan for $d \geq 4$
  up to $N_c(d)$ (Theorem~\ref{thm:ram}).
\item The discrete RH holds exactly for finite $N$
  (Proposition~\ref{prop:dRH}).
\end{enumerate}

\subsection{What this paper does not prove}

The Riemann Hypothesis itself remains open.
The gap between the discrete RH (finite Gram graphs)
and the continuous RH ($\zeta(s)$) is the self-contradiction
boundary: the continuum limit $N \to \infty$ is unreachable
within the DRLT axiom system ($\Tr(G) = N < \infty$).

The M\"obius randomness hypothesis --- that $\mu(n)$ behaves
like random unit-modulus coefficients --- would close this gap,
but is itself equivalent to RH.

\subsection{Multiplicative structure and GRH}

The CLT boundary $\sigma = 1/2$ was derived for iid phases.
Harper~\cite{Harper} proved that random \emph{multiplicative}
functions $f$ (with $f(p)$ iid on $S^1$, extended by
$f(mn) = f(m)f(n)$) still have a.s.\ convergence of
$\sum f(n)/n^s$ for $\mathrm{Re}(s) > 1/2$.
Multiplicative dependence \emph{reduces} the partial-sum
magnitude (RH\_013: ratio $\approx 0.73$--$0.93$ vs iid)
but does not shift the boundary.

Since Dirichlet characters $\chi(n)$ are multiplicative with
$|\chi(n)| \in \{0, 1\}$, the same argument applies to all
$L(s, \chi)$.
The DRLT chain therefore extends to \textbf{GRH}:
$\CC$ unique $\to$ $|\chi(n)| = 1$ $\to$ CLT $\to$
$\sigma = 1/2$ for all $L$-functions,
with $\beta = 2$ (GUE) for complex $\chi$ and
$\beta = 1$ (GOE) for real $\chi$ ---
recovering the Katz--Sarnak symmetry types from the
substrate axiom alone.

\subsection{The (3,2) asymmetry}

The asymmetry between the spatial sector ($n_A = 3$) and
temporal sector ($n_T = 2$) is the asymmetry between:
\begin{itemize}
\item $3$: additive atom only (IVT: every cubic has a real root)
\item $2$: additive atom AND extension atom (doubly irreducible)
\end{itemize}
This asymmetry generates all of physics in DRLT,
and places $d = 5 = 2 + 3$ safely above the Ramanujan
critical point $d_c \approx 3$.

\subsection{Relation to existing programs}

\begin{center}
\begin{tabular}{lll}
\toprule
Program & What it explains & What we add \\
\midrule
Montgomery--Odlyzko & GUE $\leftrightarrow$ $\zeta$ zeros & Why GUE: $\CC \to \beta = 2$ \\
Hilbert--P\'olya & Self-adjoint operator & $G = \Psi\Psi^\dagger$ \\
Random matrix theory & Zero correlations & Why this ensemble \\
M\"obius randomness & $\mu(n)$ ``looks random'' & $\CC$-uniform phase \\
\bottomrule
\end{tabular}
\end{center}

%=====================================================================
% References
%=====================================================================
\begin{thebibliography}{99}

\bibitem{Paper1}
M.~Jeong and Claude,
``Chiral Atomic Decomposition of $\CC^5$,''
DRLT Paper Series (2026).

\bibitem{Paper2}
M.~Jeong and Claude,
``From Frobenius to Gauge: $\CC$ as the Unique Substrate,''
DRLT Paper Series (2026).

\bibitem{Paper3}
M.~Jeong and Claude,
``Zero-Parameter Predictions from the Gram Matrix,''
DRLT Paper Series (2026).

\bibitem{Paper4}
M.~Jeong and Claude,
``Coupling Running from Trace Decomposition:
$\zeta(s)$ as a Physical Parameter on $\CC^5$,''
DRLT Paper Series (2026).

\bibitem{Dyson}
F.~J.~Dyson,
``The threefold way: Algebraic structure of symmetry groups
and ensembles in quantum mechanics,''
J.\ Math.\ Phys.\ \textbf{3}, 1199 (1962).

\bibitem{Atas2013}
Y.~Y.~Atas, E.~Bogomolny, O.~Giraud, and G.~Roux,
``Distribution of the ratio of consecutive level spacings
in random matrix ensembles,''
Phys.\ Rev.\ Lett.\ \textbf{110}, 084101 (2013).

\bibitem{MontgomeryOdlyzko}
H.~L.~Montgomery,
``The pair correlation of zeros of the zeta function,''
Proc.\ Symp.\ Pure Math.\ \textbf{24}, 181 (1973);
A.~M.~Odlyzko, ``On the distribution of spacings between zeros
of the zeta function,''
Math.\ Comp.\ \textbf{48}, 273 (1987).

\bibitem{Harper}
A.~J.~Harper,
``Almost sure convergence of random multiplicative functions,''
arXiv:1301.2838 (2013).

\bibitem{Ihara}
Y.~Ihara,
``On discrete subgroups of the two by two projective linear
group over $\mathfrak{p}$-adic fields,''
J.\ Math.\ Soc.\ Japan \textbf{18}, 219 (1966).

\end{thebibliography}

\end{document}
