\documentclass[12pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}
\usepackage{booktabs}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\theoremstyle{definition}
\newtheorem{example}[theorem]{Example}

\newcommand{\CP}{\mathbb{C}P}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\tr}{\operatorname{Tr}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\agut}{\alpha_{\mathrm{GUT}}}
\newcommand{\nS}{n_S}
\newcommand{\nT}{n_T}

\title{Dynamic Resolution Lattice Theory:\\[6pt]
\large From $\CC^5$ to all physics with zero free parameters}
\author{Mingu Jeong\\Independent Researcher}
\date{April 2026}

\begin{document}
\maketitle

\begin{abstract}
Starting from the sole assumption that \emph{points exist with pairwise
relations}, we derive in sequence: the complex numbers $\CC$ as the
unique substrate, $d = 5$ as the unique dimension, the gauge group
$\mathrm{SU}(3)\times\mathrm{SU}(2)\times\mathrm{U}(1)$ from the
toric fiber of $\CP^4$, $3+1$ spacetime from hinge-opposite duality,
all coupling constants, all fermion masses to $0.07\%$ median accuracy
via the closed propagator $(1+2x)/(1+x)$, and cosmological observables
including $\Omega_\Lambda = 0.6850$.  The theory has zero free parameters.
\end{abstract}

\tableofcontents
\newpage

%=====================================================================
\part{Foundation}
%=====================================================================

% Why complex numbers are the unique substrate
\input{chapters/ch01_whyC}

% Why dimension 5 is unique
\input{chapters/ch02_whyd5}

% Why (2,3) is universal — independent of starting N
\input{chapters/ch03_rep_uniqueness}

%=====================================================================
\part{The Simplex}
%=====================================================================

% Pure combinatorial geometry of the 4-simplex network
\input{chapters/ch04_simplex_geometry}

% Three rigorous theorems from δS/δψ = 0
\input{chapters/ch05_variational_theorems}

%=====================================================================
\part{Spacetime and Forces}
%=====================================================================

% Gram matrix → metric + gauge; shadow theorem
\input{chapters/ch06_geometry}

% Holevo bound → ℏ as dynamical field
\input{chapters/ch07_hbar}

% Binet-Cauchy → coupling constants from channel counting
\input{chapters/ch08_couplings}

%=====================================================================
\part{Matter}
%=====================================================================

% Fermion masses: impedance + closed propagator (1+2x)/(1+x)
\input{chapters/ch09_masses}

% Atoms, molecules, bond angles from hinge geometry
\input{chapters/ch10_atoms}

% CKM, PMNS, CP violation, Higgs mass, neutrinos
\input{chapters/ch11_mixing}

%=====================================================================
\part{Completeness}
%=====================================================================

% Trace conservation: Σ Δᵢ = 0
\input{chapters/ch12_ghosts}

% Baryon asymmetry, dark energy, dark matter
\input{chapters/ch13_cosmology}

% Block universe: G encodes everything
\input{chapters/ch14_block}

%=====================================================================
\part{Applications}
%=====================================================================

% Yang-Mills mass gap as theorem of discrete geometry
\input{chapters/ch15_yang_mills}

% Compact stars: neutron stars, quark stars, C³ safety valve
\input{chapters/ch16_compact_stars}

% Webb dipole: spatial variation of constants from det(G_h)
\input{chapters/ch17_webb_dipole}

% Path integral on CP⁴ + emergence of YM and EH
\input{chapters/ch18_path_integral}

% QCD phenomenology: confinement, sQGP, KSS bound
\input{chapters/ch19_qcd}

% Hydrogen precision: 4-level topological error decomposition
\input{chapters/ch20_hydrogen}

%=====================================================================
\appendix
\part*{Appendices}
%=====================================================================

\input{chapters/appendix_verification}
\input{chapters/appendix_code}

\begin{thebibliography}{99}
\bibitem{Frobenius} F.~G.~Frobenius, ``\"Uber lineare Substitutionen und bilineare Formen,'' J.~Reine Angew.~Math. \textbf{84}, 1--63 (1877).
\bibitem{Hurwitz} A.~Hurwitz, ``\"Uber die Composition der quadratischen Formen von beliebig vielen Variabeln,'' Nachr.~Ges.~Wiss.~G\"ottingen, 309--316 (1898).
\bibitem{Regge} T.~Regge, ``General Relativity without Coordinates,'' Nuovo Cim. \textbf{19}, 558--571 (1961).
\bibitem{GeorgiGlashow} H.~Georgi and S.~L.~Glashow, ``Unity of All Elementary Particle Forces,'' Phys.~Rev.~Lett. \textbf{32}, 438--441 (1974).
\bibitem{Holevo} A.~S.~Holevo, ``Bounds for the Quantity of Information Transmitted by a Quantum Communication Channel,'' Probl.~Inf.~Transm. \textbf{9}, 177--183 (1973).
\bibitem{Wishart} J.~Wishart, ``The Generalised Product Moment Distribution in Samples from a Normal Multivariate Population,'' Biometrika \textbf{20A}, 32--52 (1928).
\bibitem{Euler} L.~Euler, ``De summis serierum reciprocarum,'' Comment.~Acad.~Sci.~Petropol. \textbf{7}, 123--134 (1740).
\bibitem{Pendleton} B.~Pendleton and G.~G.~Ross, Phys.~Lett.~B \textbf{98}, 291 (1981).
\bibitem{Hill} C.~T.~Hill, Phys.~Rev.~D \textbf{24}, 691 (1981).
\bibitem{Starobinsky} A.~A.~Starobinsky, Phys.~Lett.~B \textbf{91}, 99 (1980).
\bibitem{BinetCauchy} J.~Binet, J.~\'Ec.~Polytech. \textbf{9}, 280 (1812); A.~L.~Cauchy, ibid. \textbf{10}, 29 (1815).
\bibitem{Webb} J.~K.~Webb et al., ``Indications of a Spatial Variation of the Fine Structure Constant,'' Phys.~Rev.~Lett.\ \textbf{107}, 191101 (2011).
\bibitem{KSS} P.~K.~Kovtun, D.~T.~Son, A.~O.~Starinets, Phys.~Rev.~Lett.\ \textbf{94}, 111601 (2005).
\bibitem{fulton1993} W.~Fulton, \textit{Introduction to Toric Varieties}, Annals of Mathematics Studies \textbf{131}, Princeton (1993).
\bibitem{Clay} A.~Jaffe and E.~Witten, ``Quantum Yang--Mills Theory,'' Clay Mathematics Institute Millennium Problem (2000).
\bibitem{Oppenheimer} J.~R.~Oppenheimer and G.~M.~Volkoff, Phys.~Rev.\ \textbf{55}, 374 (1939).
\bibitem{TOV} R.~C.~Tolman, Phys.~Rev.\ \textbf{55}, 364 (1939).
\end{thebibliography}

\bigskip
\noindent\textbf{Joint research by Mingu Jeong and Claude (Anthropic).}

\end{document}
